---
title: "Chapter 5: network-based analysis"
author: "Eli Galle, Emmeline Van Rijswijck"
date: "`r Sys.Date()`"
output: html_document
---

In this chapter, we will study the patterns of gene expression in paralogs using gene coexpression networks. We will test whether any of the duplication modes tends to create genes that are overrepresented in the network hubs.


Load packages
```{r setup, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(tidyverse)
library(here)
library(BioNERO) # BiocManager::install("almeidasilvaf/BioNERO")
library(exdiva) # BiocManager::install("almeidasilvaf/exdiva")
library(HybridExpress)
```

For making figures work on hpc
```{r}
options(bitmapType = "cairo")
```

Loading in expression and paralog data from chapter 2. Coexpression network will be constructed using pseudobulked expression data to avoid problems with sparse expression.
```{r}
sce <- readRDS(here("chapters", "summarized_sce_paralogs.rds")) 
dup_info <- readRDS(here("Duplication_data", "duplication_info.rds"))
dup_genes <- dup_info$genes
dup_pairs <- dup_info$pairs
rm(dup_info)
```


## 5.1 GCN inference

GCN inference is done using the package BioNERO. This package implements the WGCNA algorithm. 

Expression values are transformed into a pairwise similarities, calculated based on the pairwise gene correlations. An unsigned method treats negative correlations the same as positive by taking the absolute value. In a signed method the transformation is constructed a little different to only pick up on positive correlations.

Next, the similarity matrix is raised to a power beta to define the adjacency matrix. This is done to amplify distances. Values close to 1 will change relatively little while lower values will decrease. Using a higher beta results in a more scale-free topology (= most nodes have little edges, selected amount of a lot; desired characteristic in biological network). However, it also reduces mean connectivity.

The resulting network can then be analyzed using clustering methods to identify modules of genes that are adjacent.

```{r}
# to speed up calculations
# change nThreads depending on how many you have (parallel::detectCores())
WGCNA::allowWGCNAThreads(nThreads = 36)
```

GCN inference is done with log normalized counts
```{r}
mat <- assay(sce, "logcounts.sum")
```

We filter genes that are not (lowly) expressed
```{r}
as.data.frame(colData(sce)) |>
  group_by(Organ) |>
  summarise(n_celltypes = n_distinct(Celltype_original))
```
Organ with least amount of cell types has 6 (14%), want to keep organ specific genes so min_percentage_samples is set to 10%
```{r}
mat <- remove_nonexp(mat, method = "percentage", min_exp=0.01, min_percentage_samples = 0.1)
```

Optimal beta value is determined with the function SFT_fit()
(beta = controls how much you emphasize strong correlations and suppress weak ones; correlations are converted by the power of beta)

In general, (biological) networks often display a scale free topology, with many genes being lowly connected and few highly connected. This behavior is often used as quality control of networks and is measured with SFT.R.sq. This value normally increases with beta. An appropriate beta is then selected as the lowest one meeting SFT.R.sq threshold (eg 0.8), while also taking into account the connectivity (don't want it too low).

```{r}
sft <- SFT_fit(mat, net_type = "signed hybrid", cor_method = "spearman")
```

Take a look at results
```{r}
sft$power
sft$plot
```
SFT.R.sq and truncated.R.sq do not reach the threshold value, indicating that the network does not perfectly adhere to a scale free topology. A beta value of 10 is chosen as SFT.R.sq plateaus at higher values.

Infer GCN 
```{r}
run_gcn = FALSE

if (run_gcn) {
    gcn <- exp2gcn(
    mat,
    net_type = "signed hybrid", # hybrid: sim = cor if cor>0 else 0
    SFTpower = 10,       
    cor_method = "spearman",   
    return_cormat = FALSE,      # whether you want the correlation matrix
    verbose = TRUE
  )
  saveRDS(gcn, "gcn.rds")

} else {
  gcn <- readRDS(here("chapters", "gcn.rds"))
}
```

## 5.2 GCN exploration

```{r}
gcn$genes_and_modules |> nrow()
gcn$genes_and_modules$Modules |> table()
```
The paralog genes were clustered into 3 different modules (MEgrey is a trash bin module). The overwhelming majority belong to the antiquewhite2 cluster.
```{r}
plot_dendro_and_colors(gcn)
```
It is evident that a lot of initial clusters were merged, indicating they had similar module eigenvectors.

```{r}
metadata_organ <- data.frame(sub(".*_", "", colnames(mat)))
rownames(metadata_organ) <- colnames(mat)
colnames(metadata_organ) <- "Organ"
metadata_organ

MEtrait_organ <- module_trait_cor(exp = mat, MEs = gcn$MEs, metadata = metadata_organ)
plot_module_trait_cor(MEtrait_organ)
```
The antiquewhite2 module is characterized by higher expression levels in the shoot axis and relatively low expression in the flower. The darkgreen module has higher expression in the shoot axis as well, but is further characterized by relatively low expression in the root. The expression correlation across organs is less pronounced for the thistle module, with only a moderately high expression in roots. 


## 5.3 Hub genes and duplication mode
Extract hubs; groups of highly co-expressed genes
```{r}
# Extract hubs,
final_df <- cbind(gcn$genes_and_modules, gcn$kIN) |>
  filter(Modules != "grey") # grey module is the trash bin module

# split in tables for each module; want every module in the gcn to be represented
# get top 10% 
hubs <- lapply(split(final_df, final_df$Modules), function(x) {
  hubs_vec <- x |>
    arrange(-kWithin) |>
    slice_head(n = round(nrow(x) * 0.1)) |>
    pull(Genes)
  
  return(hubs_vec)
  }
)

# converting list with hub genes for each module into vector of all hub genes
# don't need to know from which module they were, so unname
hubs <- unname(unlist(hubs))
```

```{r}
cat("Connectivity of hub genes:\n")
gcn$kIN[gcn$genes_and_modules$Genes %in% hubs, ] |> pull(kWithin) |> summary() |> round(3)
cat("\n")
cat("Connectivity of non-hub genes:\n")
gcn$kIN[!gcn$genes_and_modules$Genes %in% hubs, ] |> pull(kWithin) |> summary() |> round(3)
```

Summarize number of (paralog) hub genes per duplication mode, compare with background frequency.
```{r}
# count number of (paralog) hub genes per duplication mode
dup_genes[dup_genes$gene %in% hubs, ]$type |> 
  table() |>
  prop.table() |> 
  round(3) * 100 #display as percentage

# count number of (paralog) genes per duplication mode (=background frequency)
dup_genes$type |> 
  table() |> 
  prop.table() |> 
  round(3) * 100 #display as percentage
```

Statistical test of over representation
```{r}
ora_HubsDup <- HybridExpress::ora(genes = hubs,
    annotation = dup_genes,
    background = dup_genes$gene,
    max_setsize = 100000,
    alpha = 1 # want all p-values for plot
  )
ora_HubsDup
```

Visualize with plot
```{r}
# Colors for background
# Plot background
bg <- grid::linearGradient(colorRampPalette(c("gray90", "white"))(100))

# Define color palette for duplication modes
dup_pal <- c(
    SD = "#000000",  
    TD = "#E69F00",
    PD = "#56B4E9",
    rTRD = "#009E73",
    dTRD = "#F0E442",
    DD = "#0072B2"
)
```

```{r}
# Reorder the df to the order in the color pallete for the duplication modes
ora_HubsDup <- ora_HubsDup[match(names(dup_pal), ora_HubsDup$term),]

figure_ora <- ora_HubsDup |>
  mutate(
    term = factor(term, levels = c("SD",
                                   "TD",
                                   "PD",
                                   "rTRD",
                                   "dTRD",
                                   "DD")),
    logP = -log10(padj),
    significant = ifelse(padj < 0.05, TRUE, FALSE),
    symbol = case_when(
      padj > 0.05  ~ "",
      padj > 0.01  ~ "*",
      padj > 0.001 ~ "**",
      !is.na(padj) ~ "***",
      TRUE ~ NA_character_
    )
  ) |> 
  ggplot(aes(x=genes, y=term)) +
    geom_point(
      aes(fill=term, size = logP, alpha = significant),
      color = "gray20", pch = 21
    ) +
    scale_size(range = c(4, 9)) +
    scale_alpha_manual(values = c(0.3, 1)) +
    scale_fill_manual(values = dup_pal) +
    geom_text(aes(label = symbol), vjust = -0.5, size = 5) +
    scale_x_continuous(limits = c(0, 900)) +
    theme_classic() +
    guides(fill = "none", alpha = "none") +
    labs(
        title = "Duplication mode of hub genes in coexpression networks",
        subtitle = "Bright dots with asterisks indicate significant overrepresentation; P <0.05",
        x = "Number of hub genes", y = NULL,
        size = expression(-Log[10] ~ italic(P))
    )

figure_ora
```
Paralogs produced by segmental duplication (SD) are significantly overrepresented in the hubs of the gene coexpression network.


# additional underepresentation analysis

Now we test for statistical underrepresentation
```{r}
# Custom underrepresentation test using Fisher's exact test (less alternative)
ora_under <- do.call(rbind, lapply(unique(dup_genes$type), function(mode) {
  
  in_hubs_and_mode    <- sum(hubs %in% dup_genes$gene[dup_genes$type == mode])
  in_hubs_not_mode    <- sum(hubs %in% dup_genes$gene[dup_genes$type != mode])
  not_hub_and_mode    <- sum(!dup_genes$gene[dup_genes$type == mode] %in% hubs)
  not_hub_not_mode    <- sum(!dup_genes$gene[dup_genes$type != mode] %in% hubs)
  
  mat <- matrix(c(in_hubs_and_mode, in_hubs_not_mode,
                  not_hub_and_mode, not_hub_not_mode), nrow = 2)
  
  test <- fisher.test(mat, alternative = "less")  # "less" = underrepresentation
  
  data.frame(
    term   = mode,
    genes  = in_hubs_and_mode,
    pval   = test$p.value
  )
}))

# Adjust p-values for multiple testing
ora_under$padj <- p.adjust(ora_under$pval, method = "BH")
ora_under
```
```{r}
ora_under <- ora_under[match(names(dup_pal), ora_under$term), ]

figure_ora_under <- ora_under |>
  mutate(
    term = factor(term, levels = c("SD", "TD", "PD", "rTRD", "dTRD", "DD")),
    logP = -log10(padj),
    significant = ifelse(padj < 0.05, TRUE, FALSE),
    symbol = case_when(
      padj <= 0.05 & padj > 0.01  ~ "*",
      padj <= 0.01 & padj > 0.001 ~ "**",
      padj <= 0.001                ~ "***",
      TRUE                         ~ ""
    )
  ) |>
  ggplot(aes(x = genes, y = term)) +
    geom_point(
      aes(fill = term, size = logP, alpha = significant),
      color = "gray20", pch = 21
    ) +
    scale_size(range = c(4, 9)) +
    scale_alpha_manual(values = c(0.3, 1)) +
    scale_fill_manual(values = dup_pal) +
    geom_text(aes(label = symbol), vjust = -0.5, size = 5) +
    scale_x_continuous(limits = c(0, 900)) +
    theme_classic() +
    guides(fill = "none", alpha = "none") +
    labs(
      title = "Underrepresentation of duplication modes in hub genes",
      subtitle = "Bright dots with asterisks indicate significant underrepresentation; P <0.05",
      x = "Number of hub genes", y = NULL,
      size = expression(-Log[10] ~ italic(P))
    )

figure_ora_under
```

Saving figures
```{r}
ggsave(here("plots_and_figures", "figure_ORA_DupMode_HubGenes2.png"), 
       figure_ora,
       width = 7,
       height = 5)

ggsave(here("plots_and_figures", "figure_URA_DupMode_HubGenes2.png"), 
       figure_ora,
       width = 7,
       height = 5)
```


```{r}
sessionInfo()
```

