---
title: "chapter1"
author: "Noor Ottevaere"
date: "2025-11-15"
output: html_document
---

```{r, message=FALSE}
library(zellkonverter)
library(SingleCellExperiment)
library(R.utils)
library(here)
library(tidyverse)
library(scales)
library(reticulate)
library(scuttle)
library(scater)
library(scran)
library(batchelor)
library(DropletUtils)
library(Rtsne)
library(tidyr) 
library(ggplot2) 
library(DelayedArray)
```

###################
## Read in Data ##
###################

```{r}
data_dir <- here("scRNAseq_data", "zea_mays")
```

```{r}
### Don't run this again if files are already unzipped
if (FALSE) {
  files <- list.files(path = data_dir, pattern = "\\.h5ad\\.gz$", full.names = TRUE)
  for (f in files) {
    gunzip(f, overwrite = TRUE, remove = FALSE)   # removes .gz and leaves .h5ad
    }
  }

```

```{r}
unziped_files <- list.files(path=data_dir, pattern = "\\.h5ad$", full.names = TRUE)
```

## convert to list of single cell experiments

```{r, message = FALSE}
sce_list <- lapply(unziped_files, readH5AD)
names(sce_list) <- sub("\\.h5ad$", "", basename(unziped_files))
```

##########################
## Datasets Exploration ##
##########################

```{r}
sce_info <- sapply(sce_list, function(x){
  list(
    organ = unique(colData(x)$Organ),
    tissue =  unique(colData(x)$Tissue),
    age = unique(colData(x)$ACE),
    n_experiments = length(unique(colData(x)$Orig.ident)),  # not really sure what these are (technical replicates, biological replictes, ...?)
    n_cells = ncol(x),
    n_celltypes = length(unique(colData(x)$Celltype)),
    total_genes = nrow(x),
    Avg_genes = round(mean(colData(x)$nFeature_RNA)),
    Avg_library_size = round(mean(colData(x)$nCount_RNA))
  )
})

sce_info["tissue", "SRP325657_2"] <- "Leaf abaxial mature peel" # different suffixes here were different samples (all of leaf abaxial mature peel tissues)

sce_info <- apply(sce_info, c(1,2), function(x) as.character(x)) # couldn't figure out how to unlist cells already in sce_info creation --> everything is as a character now, if needed later we can change the numbers to numeric values i think? 
sce_info <- as.data.frame(sce_info)
print(sce_info)
```
## Visualization
```{r}
# transfer into tidy format for ggplot visualization
sce_info$stat <- rownames(sce_info)
sce_info_tidy <- sce_info |>
  pivot_longer(cols = -stat, names_to = "sample", values_to = "value")
```

```{r}
sce_info_tidy %>% 
  filter(stat %in% c("n_cells", 
                     "n_celltypes", 
                     "total_genes", 
                     "Avg_genes", 
                     "Avg_library_size", 
                     "n_experiments")) %>%
  ggplot(aes(x = sample, y = as.numeric(value))) +
  geom_col() +
  facet_wrap(~stat, scales = "free_y") +
  scale_y_continuous(breaks = pretty_breaks(n = 4)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.3)) +
  labs(x = NULL, y = NULL)
```

Check annotated celltypes in datasets
```{r}
cell_types <- sapply(sce_list, function(x){
  list(
    c(unique(colData(x)$Celltype), rep( "/", 13 - length(unique(colData(x)$Celltype)) ))
  )
})
cell_types <- as.data.frame(cell_types)
cell_types
```
Problems
- contamination and unknown
- SRP145013:
    - root pith, xylem part of root stele
- SRP224648 and SRP281914: only 2 celltypes; not really enough to define REB
  --> dont use for individual dataset analysis (but do use for total)
- SRP325657_1: only 3 celltypes
  --> enough for individual analysis
- SRP325657_2: Pavement cell-N and Pavement cell-A	 
    PCs-A: PCs with active anthocyanin biosynthesis
    PCs-N: PCs lacking active anthocyanin biosynthesis
    natural celltypes (not transformed)
- SRP417893: leaf rim; dubious celltype
    "Here we use single-cell RNA-sequencing analyses to identify a ‘rim’ cell type 
    present at the margins of maize leaf primordia. Cells in the leaf rim have a 
    distinctive identity and share transcriptional signatures with proliferating 
    ligule cells"
    can probably be seen as valid celltype, otherwise merged with leaf primordium
- resolution problem; 
  some datasets have more celltypes simply because higher resolution or more 
  similar celltypes
  eg SRP272727_23_26; multiple meristem subcategories
  eg SRP266308_1; vascular tissue <-> other; xylem & phloem
  -->
  
UMAP plot of each data set colored by cell type
```{r}
for (i in 1:length(sce_list)) {
  plot <- plotReducedDim(sce_list[[i]], dimred = "X_umap", colour_by = "Celltype") +
    ggtitle(names(sce_list[i]))
  print(plot)
} 
```

Violin plot of number of expressed genes per dataset
```{r}
df <- map_df(names(sce_list), function(name) {
  sce <- sce_list[[name]]
  expr <- assay(sce, "X")   # your assay

  tibble(
    dataset = name,
    n_genes = colSums(expr > 0),
    tissue = colData(sce)$Tissue  # automatically gets tissue
  )
})

ggplot(df, aes(x = dataset, y = n_genes, fill = tissue)) +
  geom_violin(trim = FALSE) +
  scale_fill_manual(values= palette.colors()) +
  scale_y_log10() +
  coord_flip() +
  labs(y = "Number of expressed genes") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
  )
```
-> We see that for example the leaf datasets are comparable in expressed genes even though they are two different datasets
add more descirptive title,

#####################
## Quality control ##
#####################

## Quality control of datasets

## filtering of bad genotypes, celltypes and conditions
```{r}
bad_celltypes <- c("Unknow", 
                   "Contamination")

bad_genotypes <- c("B73 (ns)") 

bad_conditions <- c("Kn1-O/+",  
                    "Nitrate-")
sce_list <- lapply(sce_list, function(sce) {
  keep <- !colData(sce)$Celltype %in% bad_celltypes &
          !colData(sce)$Genotype %in% bad_genotypes &
          !colData(sce)$Condition %in% bad_conditions
  sce[, keep]
})

rm(bad_celltypes, bad_genotypes, bad_conditions)
```
- SRP417893: two genotypes: B73 and B73 (ns); only use WT
(B73 (ns): double mutations in the redundant maize Wox3 homologs narrow sheath1 (Ns1) and narrow sheath2 (Ns2) delete the marginal domain, resulting in narrow leaves that fail to form a sheathing leaf base)
- SRP266308_1: 2 conditions: Kn1-O/+  and normal cells; only use normal cells? 
Kn1-O/+ is some mutant line from what i understand from paper
- SRP335180: 2 conditions: nitrate+ and nitrate- plants 
(grown in medium with or without nitrate); only use nitrate+ since this seems the normal condition? if you leave both conditions, you can also take into account wether the plant is stressed or not, but i do not know how relevant that is for the research here and if there will be enough data to do that 


## filtering of empty droplets

```{r}
plots <- lapply(seq_along(sce_list), function(i) {
    sce <- sce_list[[i]]
    # Compute barcode ranks
    bcrank <- barcodeRanks(assay(sce, "X"))

    uniq <- !duplicated(bcrank$rank)
    # Data frame for cutoff lines
    line_df <- data.frame(
        cutoff = names(metadata(bcrank)),
        value  = unlist(metadata(bcrank))
    )

    p <- ggplot(bcrank[uniq, ], aes(rank, total)) +
        geom_point() +
        geom_hline(
            data = line_df,
            aes(color = cutoff, yintercept = value),
            lty = 2
        ) +
      

        scale_x_log10() +
        scale_y_log10() +
        labs(
            title = paste("Barcode Ranks:", names(sce_list)[i]),
            y = "Total UMI count"
        )
    p
})

# Print all plots
for (p in plots) print(p)

rm(plots, p)
```
=> filtering of empty droplets is already done

## Check if counts are integers

Mitochondrial DNA is also already filtered out (see paper scPlantDB)

```{r}
check_normalization_sparse <- function(sce) {
  
  vals <- assay(sce, "X")
  is_integer <- all(vals@x == floor(vals@x)) 
  
  # check for size factors
  sf_present <- !is.null(sizeFactors(sce))
  # output
  list(
    is_integer = is_integer,
    size_factors_present = sf_present
  )
}

# Apply to all SCEs in sce_list
norm_status <- lapply(names(sce_list), function(nm) {
  status <- check_normalization_sparse(sce_list[[nm]])
})

# Convert to data frame
norm_df <- do.call(rbind, lapply(norm_status, function(x) as.data.frame(t(x), stringsAsFactors=FALSE)))
norm_df_unlisted <- as.data.frame(apply(norm_df, c(1, 2), FUN = function(x) as.character(x)))
norm_df_unlisted

rm(norm_df, norm_df_unlisted)
```

## Mean-variance relationship

```{r}
for (name in names(sce_list)) {
  sce <- sce_list[[name]]
  
  mat <- DelayedArray(assay(sce, "X"))
  gene_means <- rowMeans(mat)
  gene_vars  <- rowVars(mat)
  
  # Free memory immediately after use
  rm(mat)
  gc()
  
  df <- data.frame(mean = gene_means, variance = gene_vars)
  
  p <- ggplot(df, aes(x = mean, y = variance)) +
    geom_point(alpha = 0.6, size = 0.7) +
    ggtitle(paste("Mean–Variance relationship:", name)) +
    scale_y_log10() + 
    scale_x_log10() +
    theme_bw()
  
  print(p)
  # Free df as well
  rm(df, gene_means, gene_vars, p)
}
```

=> Data is not logtransformed/normalised yet 
#####
## Normalisation based on quickclustering
#####

Quickcluster function:
param method = String specifying the clustering method to use. "hclust" uses hierarchical clustering while "igraph" uses graph-based clustering.

Do quickcluster for all singleCellExperiments then compute sizefactors
```{r}
set.seed(42)
sce_list <- lapply(sce_list, function(sce) {
  cl <- quickCluster(sce, assay.type = "X")
  sce <- computeSumFactors(sce, clusters = cl, assay.type = "X")
  
  plot(librarySizeFactors(sce, assay.type = "X"), sizeFactors(sce), xlab="Library size factor",
       ylab="Deconvolution size factor", log='xy', pch=16,
       col=as.integer(factor(cl)))
  abline(a=0, b=1, col="red")
  
  sce
})
```

```{r}
sce_list <- lapply(sce_list, function(sce) {
  logNormCounts(sce, assay.type="X")
})
```


```{r}
saveRDS(sce_list, file = "normalised_sce_list_quickcluster.rds")
```

```{r}
for (name in names(sce_list)) {
  sce <- sce_list[[name]]
  dec <- modelGeneVar(sce)
  varfit <- metadata(dec)
  plot(varfit$mean, varfit$var, 
       xlab="Mean of log-expression",
      ylab="Variance of log-expression")
  curve(varfit$trend(x), col="dodgerblue", add=TRUE, lwd=2)
  rm(sce, dec, varfit)
  }
```

```{r}
sce_list <- readRDS("normalised_sce_list.rds")
assays(sce_list[[1]])
```
#####
### normalisation based on celltype
#####
```{r}
set.seed(42)
sce_list <- lapply(sce_list, function(sce) {
  cl <- sce$Celltype
  sce <- computeSumFactors(sce, clusters = cl, assay.type = "X")
  
  plot(librarySizeFactors(sce, assay.type = "X"), sizeFactors(sce), xlab="Library size factor",
       ylab="Deconvolution size factor", log='xy', pch=16,
       col=as.integer(factor(cl)))
  abline(a=0, b=1, col="red")
  
  sce
})
```
```{r}
sce_list <- lapply(sce_list, function(sce) {
  logNormCounts(sce, assay.type="X")
})
```


```{r}
saveRDS(sce_list, file = "normalised_sce_list_celltype.rds")
```

```{r}
for (name in names(sce_list)) {
  sce <- sce_list[[name]]
  dec <- modelGeneVar(sce)
  varfit <- metadata(dec)
  plot(varfit$mean, varfit$var, 
       xlab="Mean of log-expression",
      ylab="Variance of log-expression")
  curve(varfit$trend(x), col="dodgerblue", add=TRUE, lwd=2)
  rm(sce, dec, varfit)
  }
```

####################
## Data Wrangling ##
####################

## Combining seperate SingleCellExperiments into 1 object

First we verify if single cell experiments can be merged without problems
(cbind attempts to combine internal metadata tables of slots in SCE object --> these slots need to have same items)
```{r}
# Verify if gene names align
print("gene names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(rownames(x), rownames(sce_list$SRP145013))))

# Verify if names in each slot of sce align
  # assays
print("assays names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(assays(x)), names(assays(sce_list$SRP145013)))))
  # colData
print("colData names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(assays(x)), names(assays(sce_list$SRP145013)))))
  # rowData
print("rowData names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(rowData(x)), names(rowData(sce_list$SRP145013)))))
  # reducedDims
print("reducedDims names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(reducedDims(x)), names(reducedDims(sce_list$SRP145013)))))
  # int_colData
print("int_colData names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(int_colData(x)), names(int_colData(sce_list$SRP145013)))))
```
reduced dimensions do not aling for all datasets
```{r}
sapply(X = sce_list, 
       FUN = function(x) names(reducedDims(x)))
```
Dataset SRP224648 has X_tsne in addition to X_umap in reduced dimension slot

Remove tsne from Dataset SRP224648
```{r}
# back up tsne in separate variable for later if needed
tsne_SRP224648 <- reducedDims(sce_list$SRP224648)$X_tsne

reducedDims(sce_list$SRP224648)$X_tsne <- NULL
names(reducedDims(sce_list$SRP224648))
```
Now everything aligns and sce list can be collapsed into one sce object

```{r}
# batch identifier already present in coldata()$Dataset

# merge
sce <- do.call(cbind, sce_list)
# do.call expands list into individual elements for cbind to bind together

# remove sce_list to save memory
rm(sce_list)

# more meaningfull name for assay
names(assays(sce)) <- "logcounts"
```


## batch effect 

Visualization first
```{r}
# Feature selection for dimensionality reduction: want genes that are 
# informative of the biological structure of the data
dec <- modelGeneVar(sce, block = sce$Dataset) # decomposition of variance
chosen.hvgs <- dec$bio > 0                    # select genes with biological variance among datasets

# saving because runTSNE takes long
if (FALSE) {
  sce <- runPCA(sce, subset_row = chosen.hvgs, ntop = 1000)
  sce <- runTSNE(sce, dimred = "PCA")
  saveRDS(sce, file = "sce_RedDim.rds")
} else {
  sce <- readRDS("sce_RedDim.rds")
}

sce$Dataset = as.factor(sce$Dataset)
```

making dataframe for TSNE plots
```{r}
color_vec <- c("#5A5156", "#E4E1E3", "#FBE426", "#F6222E", "#FE00FA", "#16FF32", "#3283FE", 
               "#FEAF16", "#B00068", "#1CFFCE", "#90AD1C", "#2ED9FF", "#DEA0FD", 
               "#AA0DFE", "#F8A19F", "#325A9B", "#C4451C", "#1C8356", "#85660D", 
               "#B10DA1", "#3B00FB", "#1CBE4F", "#FA0087", "#333333", "#F7E1A0", 
               "#C075A6", "#AAF400", "#BDCDFF", "#822E1C", "#B5EFB5", 
               "#7ED7D1", "#1C7F93", "#D85FF7", "#683B79", "#66B0FF")
color_vec36 <- c("#5A5156", "#E4E1E3", "#FBE426", "#F6222E", "#FE00FA", "#16FF32", "#3283FE", 
               "#FEAF16", "#B00068", "#1CFFCE", "#90AD1C", "#2ED9FF", "#DEA0FD", 
               "#AA0DFE", "#F8A19F", "#325A9B", "#C4451C", "#1C8356", "#85660D", 
               "#B10DA1", "#3B00FB", "#1CBE4F", "#FA0087", "#333333", "#F7E1A0", 
               "#C075A6", "#AAF400", "#BDCDFF", "#822E1C", "#B5EFB5", 
               "#7ED7D1", "#1C7F93", "#D85FF7", "#683B79","#000000", "#66B0FF")

TSNE_df <- as.data.frame(reducedDim(sce, "TSNE")) %>%  # make df from UMAP coordinates
  mutate(celltype = colData(sce)$Celltype) %>% # add celltype
  mutate(dataset = colData(sce)$Dataset) %>%   # add dataset id
  mutate(tissue = colData(sce)$Tissue)         # add tissue
```

General TSNE plot
```{r}
TSNE_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )

TSNE_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    scale_color_manual(values = color_vec36) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                     # separate aes for legend dots
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
    theme(
      legend.text = element_text(size = 8),     # textsize of legend elements
      legend.key.size = unit(0.3, "cm"))        # boxsize of legend elements
```
possible problem: shared celltypes between datasets are taken together (eg xylem). Do we want this? Otherwise a suffix of the dataset should be added to all coldata values (eg xylem in colData$Celltype--> xylem_SRP272727_23_26, xylem_SRP145013, xylem_SRP335180)

Sample effects; most variance is due to differences between datasets, not celltypes
further verify this by plotting TSNE only for leaf datasets and for shared cell types 
```{r}
TSNE_df %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) 
TSNE_df %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )
```
shared cell types between datasets dont cluster together --> strong dataset effect

same for the two root samples
```{r}
TSNE_df %>%
  filter(str_detect(tissue, "Root")) %>%      # plot leaf datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) 
TSNE_df %>%
  filter(str_detect(tissue, "Root")) %>%      # plot leaf datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )
```

## Batch correction
```{r}
set.seed(42)
merged <- correctExperiments(
  sce,
  batch = sce$Dataset,
  assay.type = "logcounts" ## Maybe we do need a PARAM parameter? 
)
```

```{r}
merged
```
-> reconstructed assay is batch effect corrected based on dataset
## TSNE on corrected data

```{r}
pcs <- reducedDims(merged)$corrected
```

```{r}
# informative of the biological structure of the data
dec <- modelGeneVar(merged, block = merged$Dataset) # decomposition of variance
chosen.hvgs <- dec$bio > 0                    # select genes with biological variance among datasets

# saving because runTSNE takes long
if (FALSE) {
  merged <- runPCA(merged, subset_row = chosen.hvgs, ntop = 1000)
  merged <- runTSNE(merged, dimred = "PCA")
  saveRDS(merged, file = "corrected_RedDim.rds")
} else {
  sce <- readRDS("corrected_RedDim.rds")
}
```


```{r}
#set.seed(42)
#tsne_out <- Rtsne(
#  pcs,
#  perplexity = 100,       
#  pca = FALSE,           
#  verbose = TRUE
#)

# Add t-SNE coordinates reduceddims
#reducedDims(merged)$TSNE <- tsne_out$Y
```
-> what is the difference between those two methods?
```{r}
TSNE_df <- data.frame(
  TSNE1 = reducedDims(merged)$TSNE[,1],
  TSNE2 = reducedDims(merged)$TSNE[,2],
  celltype = merged$Celltype, 
  dataset = merged$Dataset,
  tissue = merged$Tissue
)
```

General TSNE plot
```{r}
TSNE_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )

TSNE_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    scale_color_manual(values = color_vec36) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                     # separate aes for legend dots
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
    theme(
      legend.text = element_text(size = 8),     # textsize of legend elements
      legend.key.size = unit(0.3, "cm"))        # boxsize of legend elements
```
```{r}
TSNE_df %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) 
TSNE_df %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )
```
```{r}
TSNE_df %>%
  filter(str_detect(tissue, "Root")) %>%      # plot leaf datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) 
TSNE_df %>%
  filter(str_detect(tissue, "Root")) %>%      # plot leaf datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )
```
TO DO: 

- Check if normalisation is still needed
- check batch effects
- make matrix aggregating counts per celltype by mean, median, sum
- investigate what n_experiments is (see line 56), if these are multiple samples within each dataset we might have to correct for batch effects within each dataset aswell


- explain violin plot
- additional data visualization plots?
- within dataset quality control plots (empty droplets, doublets, low quality droplets); see tutorial or stat genomics
- tidy code and comments






