---
title: "chapter1"
author: "Noor Ottevaere"
date: "2025-11-15"
output: html_document
---

```{r, message=FALSE}
library(zellkonverter)
library(SingleCellExperiment)
library(R.utils)
library(here)
library(tidyverse)
library(scales)
```

###################
## Read in Data ##
###################

```{r}
data_dir <- here("scRNAseq_data", "zea_mays")
```

```{r}
### Don't run this again if files are already unzipped
files <- list.files(path = data_dir, pattern = "\\.h5ad\\.gz$", full.names = TRUE)
for (f in files) {
  gunzip(f, overwrite = TRUE, remove = FALSE)   # removes .gz and leaves .h5ad
}
```

```{r}
unziped_files <- list.files(path=data_dir, pattern = "\\.h5ad$", full.names = TRUE)
```

## convert to single cell experiments

```{r, message = FALSE}
sce_list <- lapply(unziped_files, readH5AD)
names(sce_list) <- sub("\\.h5ad$", "", basename(unziped_files))
```
=> sce_list is now a list of single cell experiments


##########################
## Datasets Exploration ##
##########################

```{r}
sce_info <- sapply(sce_list, function(x){
  list(
    organ = unique(colData(x)$Organ),
    tissue =  unique(colData(x)$Tissue),
    age = unique(colData(x)$ACE),
    n_experiments = length(unique(colData(x)$Orig.ident)),  # not really sure what these are (technical replicates, biological replictes, ...?)
    n_cells = ncol(x),
    n_celltypes = length(unique(colData(x)$Celltype)),
    total_genes = nrow(x),
    Avg_genes = round(mean(colData(x)$nFeature_RNA)),
    Avg_library_size = round(mean(colData(x)$nCount_RNA))
  )
})

sce_info["tissue", "SRP325657_2"] <- "Leaf abaxial mature peel" # different suffixes here were different samples (all of leaf abaxial mature peel tissues)

sce_info <- apply(sce_info, c(1,2), function(x) unlist(x)) # couldn't figure out how to unlist cells already in sce_info creation
sce_info <- as.data.frame(sce_info)
print(sce_info)
```
Visualization
```{r}
# transfer into tidy format for ggplot visualization
sce_info$stat <- rownames(sce_info)
sce_info_tidy <- sce_info %>%
  pivot_longer(cols = -stat, names_to = "sample", values_to = "value")
```

```{r}
sce_info_tidy %>% 
  filter(stat %in% c("n_cells", 
                     "n_celltypes", 
                     "total_genes", 
                     "Avg_genes", 
                     "Avg_library_size")) %>%
  ggplot(aes(x = sample, y = as.numeric(value))) +
  geom_col() +
  facet_wrap(~stat, scales = "free_y") +
  scale_y_continuous(breaks = pretty_breaks(n = 4)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.3)) +
  labs(x = NULL, y = NULL)
```

Check annotated celltypes in datasets
```{r}
cell_types <- sapply(sce_list, function(x){
  list(
    c(unique(colData(x)$Celltype), rep( "/", 13 - length(unique(colData(x)$Celltype)) ))
  )
})
cell_types <- as.data.frame(cell_types)
```
Problems
- contamination and unknow
- SRP145013:
    - root pith, xylem part of root stele
- SRP224648 and SRP281914: only 2 celltypes; not really enough to define REB
  --> dont use for individual dataset analysis (but do use for total)
- SRP325657_1: only 3 celltypes
  --> enough for individual analysis
- SRP325657_2: Pavement cell-N and Pavement cell-A	 
    PCs-A: PCs with active anthocyanin biosynthesis
    PCs-N: PCs lacking active anthocyanin biosynthesis
    natural celltypes (not transformed)
- SRP417893: leaf rim; dubious celltype
    "Here we use single-cell RNA-sequencing analyses to identify a ‘rim’ cell type 
    present at the margins of maize leaf primordia. Cells in the leaf rim have a 
    distinctive identity and share transcriptional signatures with proliferating 
    ligule cells"
    can probably be seen as valid celltype, otherwise merged with leaf primordium
- resolution problem; 
  some datasets have more celltypes simply because higher resolution or more 
  similar celltypes
  eg SRP272727_23_26; multiple meristem subcategories
  eg SRP266308_1; vascular tissue <-> other; xylem & phloem
  -->



filtering
```{r}
bad_celltypes <- c("Unknow", 
                   "Contamination")

bad_genotypes <- c("B73 (ns)") 

bad_conditions <- c("Kn1-O/+",  
                    "Nitrate-")
sce_list <- lapply(sce_list, function(sce) {
  keep <- !colData(sce)$Celltype %in% bad_celltypes &
          !colData(sce)$Genotype %in% bad_genotypes &
          !colData(sce)$Condition %in% bad_conditions
  sce[, keep]
})
```

```{r}
df <- map_df(names(sce_list), function(name) {
  sce <- sce_list[[name]]
  expr <- assay(sce, "X")   # your assay

  tibble(
    dataset = name,
    n_genes = colSums(expr > 0),
    #tissue = tissue_lookup[name]  # automatically gets tissue
  )
})

ggplot(df, aes(x = dataset, y = n_genes, fill = dataset)) +
  geom_violin(trim = FALSE) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
=> densities are not at all comparable, we need to do normalisation

- SRP417893: two genotypes: B73 and B73 (ns); only use WT
(B73 (ns): double mutations in the redundant maize Wox3 homologs narrow sheath1 (Ns1) and narrow sheath2 (Ns2) delete the marginal domain, resulting in narrow leaves that fail to form a sheathing leaf base)
- SRP266308_1: 2 conditions: Kn1-O/+  and normal cells; only use normal cells?
Kn1-O/+ is some mutant line from what i understand from paper
- SRP335180: 2 conditions: nitrate+ and nitrate- plants 
(grown in medium with or without nitrate); only use nitrate+ since this seems the normal condition?



TO DO: 
- filter datasets: already done on celltype, genotype and condition: check whether there's additional stuff that needs to be thrown out
- integrate all data sets into one `SingleCellExperiment` object while taking potential batch effects into account
- make matrix aggregating counts per celltype by mean, median, sum

- put unzip block into if statement
- investigate what n_experiments is (see line 56)






