---
title: 'Chapter 2: Data summary'
author: "Eli Galle"
date: "2026-02-20"
output: html_document
---

In this chapter we will summarize the processed data from chapter 1, enabling further analysis.

Loading libraries
```{r}
library(SingleCellExperiment)
library(scuttle)
library(stringr)
```


As the quality of batch correction was not satisfactory, we will work further with uncorrected counts. This will have to be taken into account in subsequent analysis.
```{r}
sce <- readRDS("uncorrected_sce_RedDim.rds")
```

First we will make a new colData variable indicating the cell type and tissue combination for each cell. This is the variable across we will summarize our counts


```{r}
# string manipulation to make resulting variable more readable
# eg "RootCap_RootTip" in stead of "Root cap_Root tip"

celltypes <- sce$Celltype %>%
  str_to_title() %>%
  str_remove_all(" ")

tissues <- sce$Tissue %>%
  str_to_title() %>%
  str_remove_all(" ")

# create new colData variable
sce$Celltype_Tissue <- paste(celltypes, tissues, sep="_")

rm(celltypes, tissues)
```

Summarize/aggregate counts across Celltype_Tissue
```{r}
summarized_sce <- aggregateAcrossCells(
  sce, 
  ids = sce$Celltype_Tissue,
  use.assay.type = c("counts","logcounts"),
  statistics = c("sum", "mean", "median"),
  store.number = "ncells")                   # store number of cells 
```




QUESTIONS:
??? not sure whether to summarize across Celltype_Tissue, _Dataset or _Organ ??? (line 23-40)













