---
title: "Chapter 1: Data Prepreration"
author: "Noor Ottevaere"
date: "2025-11-15"
output: html_document
---

```{r, message=FALSE}
library(zellkonverter)
library(SingleCellExperiment)
library(R.utils)
library(here)
library(tidyverse)
library(scales)
library(reticulate)
library(scuttle)
library(scater)
library(scran)
library(batchelor)
library(DropletUtils)
library(Rtsne)
library(tidyr) 
library(ggplot2) 
library(DelayedArray)
library(ragg) #to save plots 
library(scater)
library(PCAtools)
```

##################
## Read in Data ##
##################

Setting root directory to files
```{r}
data_dir <- here("scRNAseq_data", "zea_mays")
```

Getting file directories
```{r}
files <- list.files(path=data_dir, pattern = "\\.rds$", full.names = TRUE)
```

Reading in RDS files into list of single cell experiments
```{r}
sce_list <- lapply(files, readRDS)
names(sce_list) <- sub("\\.rds$", "", basename(files))
```


##########################
## Datasets Exploration ##
##########################

Gathering info on the datasets
```{r}
sce_info <- sapply(sce_list, function(x){
  list(
    organ = unique(colData(x)$Organ),
    tissue =  unique(colData(x)$Tissue),
    age = unique(colData(x)$ACE),
    n_experiments = length(unique(colData(x)$Orig.ident)),  # not really sure what these are (technical replicates, biological replicates, ...?)
    n_cells = ncol(x),
    n_celltypes = length(unique(colData(x)$Celltype)),
    total_genes = nrow(x),
    Avg_genes = round(mean(colData(x)$nFeature_RNA)),
    Avg_library_size = round(mean(colData(x)$nCount_RNA))
  )
})

sce_info["tissue", "SRP325657_2"] <- "Leaf abaxial mature peel" # different suffixes here were different samples (all of leaf abaxial mature peel tissues)

sce_info <- apply(sce_info, c(1,2), function(x) as.character(x)) 
sce_info <- as.data.frame(sce_info)
print(sce_info)
```

Visualization
```{r}
# transfer into tidy format for ggplot visualization
sce_info$stat <- rownames(sce_info)
sce_info_tidy <- sce_info |>
  pivot_longer(cols = -stat, names_to = "sample", values_to = "value")
```

```{r}
tissue_map <- sce_info_tidy |>
  filter(stat == "tissue") |>
  select(sample, tissue = value)
```

```{r}
sce_info_tidy2 <- sce_info_tidy |>
  left_join(tissue_map, by = "sample")
```

```{r}
sce_info_plot <- sce_info_tidy2 %>%
  filter(stat %in% c("n_cells", 
                     "n_celltypes", 
                     "total_genes", 
                     "Avg_genes", 
                     "Avg_library_size", 
                     "n_experiments")) %>%
  ggplot(aes(x = sample, y = as.numeric(value))) +
  geom_col(fill = "steelblue") +
  facet_wrap(~stat, scales = "free_y") +
  scale_x_discrete(labels = sce_info_tidy2$tissue) +
  scale_y_continuous(breaks = pretty_breaks(n = 4)) +
  labs(x = NULL, y = NULL) +
  theme(
      plot.title = element_text(face = "bold", size = 24),
      axis.text.x = element_text(angle = 90, vjust = 0.3),
      panel.grid = element_blank(),
      axis.line = element_blank(),
      axis.ticks = element_blank(),
  ) 
sce_info_plot
```
Save the plots and figures in a folder 
```{r}
if (FALSE) {
  out_dir <- here("plots_and_figures")
  outfile <- file.path(out_dir, paste0("plot_", 'sce_info_',".png")) #to overwrite an other plot with the same name 
  ggsave(filename = outfile, plot = sce_info_plot)
  }
```
Check annotated celltypes in datasets
```{r}
cell_types <- sapply(sce_list, function(x){
  list(
    c(unique(colData(x)$Celltype), rep( "/", 13 - length(unique(colData(x)$Celltype)) ))
  )
})
cell_types <- as.data.frame(cell_types)
cell_types
```
Possible problems to keep in mind for downstream analysis:
- contamination and unknown
  --> both removed
- SRP224648 and SRP281914: only 2 celltypes; not really enough to define REB
  --> dont use for individual dataset analysis (but do use for total)
- SRP325657_1: only 3 celltypes
  --> enough for individual analysis
- SRP325657_2: Pavement cell-N and Pavement cell-A	 
    PCs-A: PCs with active anthocyanin biosynthesis
    PCs-N: PCs lacking active anthocyanin biosynthesis
    natural celltypes (not transformed), so both can be used
- SRP417893: leaf rim; dubious celltype
    "Here we use single-cell RNA-sequencing analyses to identify a ‘rim’ cell type 
    present at the margins of maize leaf primordia. Cells in the leaf rim have a 
    distinctive identity and share transcriptional signatures with proliferating 
    ligule cells"
    can probably be seen as valid celltype, otherwise merged with leaf primordium
- resolution problem; 
  some datasets have more celltypes simply because higher resolution or more 
  similar celltypes
  eg SRP272727_23_26; multiple meristem subcategories
  eg SRP266308_1; vascular tissue <-> other; xylem & phloem
  -->


UMAP plot of each data set colored by cell type
```{r}
for (i in 1:length(sce_list)) {
  plot <- plotReducedDim(sce_list[[i]], dimred = "UMAP", colour_by = "Celltype") +
    ggtitle(paste(names(sce_list[i])), sce_info["tissue", i])
  print(plot)
  #save plots
  if (FALSE) {
    sample_name <- names(sce_list[i])
    outfile <- file.path(out_dir, paste0("UMAP_", sample_name, ".png"))
    agg_png(outfile, width = 1600, height = 1200, res = 200)
    print(plot)  
    dev.off()
    }
} 
```
PCA plot of each data set colored by celltype
```{r}
for (i in 1:length(sce_list)) {
  plot <- plotReducedDim(sce_list[[i]], dimred = "PCA", colour_by = "Celltype") +
    ggtitle(paste(names(sce_list[i])), sce_info["tissue", i])
  print(plot)
  #save plots
  if (FALSE) {
    sample_name <- names(sce_list[i])
    outfile <- file.path(out_dir, paste0("PCA_", sample_name, ".png"))
    agg_png(outfile, width = 1600, height = 1200, res = 200)
    print(plot)  
    dev.off()
    }
} 
```

Violin plot of number of expressed genes per dataset
```{r, fig.width=7, fig.height=8}
df <- map_df(names(sce_list), function(name) {
  sce <- sce_list[[name]]
  expr <- assay(sce, "counts")   # your assay

  tibble(
    dataset = name,
    n_genes = colSums(expr > 0),
    tissue = colData(sce)$Tissue  # automatically gets tissue
  )
})

violin <- ggplot(df, aes(x = dataset, y = n_genes, fill = tissue)) +
  geom_violin(trim = FALSE) +
  scale_fill_manual(values= palette.colors()) +
  scale_y_log10() +
  coord_flip() +
  labs(y = "Number of expressed genes") +
  theme_bw() +
  ggtitle("Distribution of number of expressed genes for each dataset") +
  theme(
    panel.grid = element_blank(),
  )
plot(violin)

if (FALSE){
  outfile <- file.path(out_dir, paste0("Violinplot",".png"))
  ggsave(outfile, violin)
  }
```
We see that for example the leaf datasets are comparable in expressed genes even though they are two different datasets


#####################
## Quality control ##
#####################

## Filtering of unwanted cells 

Total number of cells before filtering
```{r}
sum(as.numeric(sce_info["n_cells",-ncol(sce_info)]))
```

Look at present genotypes
```{r}
print("Genotypes")
for (i in 1:9) {
  cat(names(sce_list)[i], ":", 
      paste(unique(colData(sce_list[[i]])$Genotype), collapse = ", "), 
      "\n", 
      sep =" ")
}
```
SRP417893: two genotypes: B73 and B73 (ns); only use WT
(B73 (ns): double mutations in the redundant maize Wox3 homologs narrow sheath1 (Ns1) and narrow sheath2 (Ns2) delete the marginal domain, resulting in narrow leaves that fail to form a sheathing leaf base)

Look at present conditions
```{r}
print("Conditions")
for (i in 1:9) {
  cat(names(sce_list)[i], ":", 
      paste(unique(colData(sce_list[[i]])$Condition), collapse = ", "), 
      "\n", 
      sep =" ")
}
```
SRP266308_1: 2 conditions: Kn1-O/+  and normal cells; only use normal cells? 
Kn1-O/+ is some mutant line from what i understand from paper

SRP335180: 2 conditions: nitrate+ and nitrate- plants 
(grown in medium with or without nitrate); only use nitrate+ since this is the normal condition.

Remove the unwanted cells
```{r}
bad_celltypes <- c("Unknow", 
                   "Contamination")

bad_genotypes <- c("B73 (ns)") 

bad_conditions <- c("Kn1-O/+",  
                    "Nitrate-")
sce_list <- lapply(sce_list, function(sce) {
  keep <- !colData(sce)$Celltype %in% bad_celltypes &
          !colData(sce)$Genotype %in% bad_genotypes &
          !colData(sce)$Condition %in% bad_conditions
  sce[, keep]
})

rm(bad_celltypes, bad_genotypes, bad_conditions)
```

```{r}
sum(as.numeric(lapply(sce_list, ncol)))
```


## Filtering of low quality cells 

Checking for empty droplets
```{r}
plots <- lapply(seq_along(sce_list), function(i) {
    sce <- sce_list[[i]]
    # Compute barcode ranks
    bcrank <- barcodeRanks(assay(sce, "counts"))

    uniq <- !duplicated(bcrank$rank)

    p <- ggplot(bcrank[uniq, ], aes(rank, total)) +
        geom_point() +
        scale_x_log10() +
        scale_y_log10() +
        labs(
            title = paste("Barcode Ranks:", names(sce_list)[i]),
            y = "Total UMI count"
        )
    p
})

# Print and save all plots
for (i in seq_along(plots)) {
  p <- plots[[i]]
  print(p)
  if (FALSE) {
    sample_name <- names(sce_list)[i]
    outfile <- file.path(out_dir, paste0("Barcode Ranks_", sample_name, ".png"))
    ggsave(outfile, p)
    }
}

rm(plots, p)
```
Filtering of empty droplets is already done
Mitochondrial DNA is also already filtered out (see paper scPlantDB)


## Normalisation and log transformation

Check if counts are integers or log transformed
```{r}
check_normalization_sparse <- function(sce) {
  
  vals <- assay(sce, "counts")
  is_integer <- all(vals@x == floor(vals@x)) 
  
  # check for size factors
  sf_present <- !is.null(sizeFactors(sce))
  # output
  list(
    is_integer = is_integer,
    size_factors_present = sf_present
  )
}

# Apply to all SCEs in sce_list
norm_status <- lapply(names(sce_list), function(nm) {
  status <- check_normalization_sparse(sce_list[[nm]])
})

# Convert to data frame
norm_df <- do.call(rbind, lapply(norm_status, function(x) as.data.frame(t(x), stringsAsFactors=FALSE)))
norm_df_unlisted <- as.data.frame(apply(norm_df, c(1, 2), FUN = function(x) as.character(x)))
norm_df_unlisted

rm(norm_df, norm_df_unlisted)
```
All datasets contain integers and don't have size factors present. 

Check if normalisation has already been done by plotting mean-variance relationship
```{r}
for (name in names(sce_list)) {
  sce <- sce_list[[name]]
  
  mat <- DelayedArray(assay(sce, "counts"))
  gene_means <- rowMeans(mat)
  gene_vars  <- rowVars(mat)
  
  # Free memory immediately after use
  rm(mat)
  gc()
  
  df <- data.frame(mean = gene_means, variance = gene_vars)
  
  p <- ggplot(df, aes(x = mean, y = variance)) +
    geom_point(alpha = 0.6, size = 0.7) +
    ggtitle(paste("Mean–Variance relationship:", name)) +
    scale_y_log10() + 
    scale_x_log10() +
    theme_bw()
  
  print(p)
  # remove df aswell
  rm(df, gene_means, gene_vars, p)
}
rm(sce)
```
Linear relation between the variance in the mean: data is not logtransformed + normalised yet. This is done ourselves in the next section 


##################################
## Within Dataset Normalisation ## 
##################################

Normalisaton is done by deconvolution:
- celltypes can have differing library sizes + compositional bias in scRNA-seq
(--> simple library size norm not valid)
- dominance of low and zero counts + many genes DE between celltypes
(--> bulk RNA norm methods problematic)

Counts are pooled from many cells (increases count size) for accurate size factor estimation. These size factors are calculated per cluster to avoid the assumption that most genes are non-DE over the entire population (only assumes that most genes are non DE within clusters). Then size factors are rescaled to be comparable across clusters.

Celltypes can be taken as clusters or clusters can be decided with Quickcluster function:

!don't run again, read in rds at the end!

## Normalisation with clusters from Quickcluster

Quickcluster function:
param method = String specifying the clustering method to use. "hclust" uses hierarchical clustering while "igraph" uses graph-based clustering.

Do quickcluster for all singleCellExperiments then compute sizefactors
```{r}
set.seed(42)
if (FALSE) {
  sce_list <- lapply(sce_list, function(sce) {
    cl <- quickCluster(sce, assay.type = "counts")
    sce <- computeSumFactors(sce, clusters = cl, assay.type = "counts")
    plot(
      librarySizeFactors(sce, assay.type = "counts"),
      sizeFactors(sce),
      xlab = "Library size factor",
      ylab = "Deconvolution size factor",
      log = "xy",
      pch = 16,
      col = as.integer(factor(cl))
    )
    abline(a = 0, b = 1, col = "red")
    sce
  })
}
```

Size factors are used to log norm transform counts
```{r}
if (FALSE) {
  sce_list <- lapply(sce_list, function(sce) {
  logNormCounts(sce, assay.type = "counts")
})

assayNames(sce_list$SRP145013)}
```

```{r}
if (FALSE) {
  saveRDS(sce_list, file = "normalised_sce_list_quickcluster.rds")
  }
```

```{r}
if (FALSE) {
  for (name in names(sce_list)) {
  sce <- sce_list[[name]]
  dec <- modelGeneVar(sce)
  varfit <- metadata(dec)
  plot(varfit$mean, varfit$var, 
      xlab="Mean of log-expression",
      ylab="Variance of log-expression")
  curve(varfit$trend(x), col="dodgerblue", add=TRUE, lwd=2)
  rm(sce, dec, varfit)
    }
  }
```

## normalisation based on celltype
Same but with annotated celltypes as clusters 

```{r}
set.seed(42)
if (FALSE){
  sce_list <- lapply(sce_list, function(sce) {
  cl <- sce$Celltype
  sce <- computeSumFactors(sce, clusters = cl, assay.type = "counts")
  
  plot(librarySizeFactors(sce, assay.type = "counts"), sizeFactors(sce),
       xlab="Library size factor",
       ylab="Deconvolution size factor", log='xy', pch=16,
       col=as.integer(factor(cl)))
  abline(a=0, b=1, col="red")
  
  sce
})}
```

```{r}
if (FALSE) {sce_list <- lapply(sce_list, function(sce) {
  logNormCounts(sce, assay.type="counts")
})}
```

```{r}
if (FALSE) {saveRDS(sce_list, file = "normalised_sce_list_celltype.rds")}
```

```{r}
if (FALSE) {for (name in names(sce_list)) {
  sce <- sce_list[[name]]
  dec <- modelGeneVar(sce)
  varfit <- metadata(dec)
  plot(varfit$mean, varfit$var, 
      xlab="Mean of log-expression",
      ylab="Variance of log-expression")
  curve(varfit$trend(x), col="dodgerblue", add=TRUE, lwd=2)
  rm(sce, dec, varfit)
  }}
```


## reading RDS

Choose Quickcluster based, this is the best --> put TRUE
```{r}
if (TRUE) {
  sce_list <- readRDS("normalised_sce_list_quickcluster.rds")
} else {
  sce_list <- readRDS("normalised_sce_list_celltype.rds")
}
```


##########################
## Data Set Aggregation ##
##########################

## Combining seperate SingleCellExperiments into 1 object

First we verify if single cell experiments can be merged without problems
(cbind attempts to combine internal metadata tables of slots in SCE object --> these slots need to have same items)
```{r}
# Verify if gene names align
print("gene names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(rownames(x), rownames(sce_list$SRP145013))))

# Verify if names in each slot of sce align
  # assays
print("assays names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(assays(x)), names(assays(sce_list$SRP145013)))))
  # colData
print("colData names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(assays(x)), names(assays(sce_list$SRP145013)))))
  # rowData
print("rowData names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(rowData(x)), names(rowData(sce_list$SRP145013)))))
  # reducedDims
print("reducedDims names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(reducedDims(x)), names(reducedDims(sce_list$SRP145013)))))
  # int_colData
print("int_colData names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(int_colData(x)), names(int_colData(sce_list$SRP145013)))))
```

reduced dimensions do not align for all data sets
```{r}
sapply(X = sce_list, 
       FUN = function(x) names(reducedDims(x)))
```
Data set SRP224648 has TSNE in addition to PCA and UMAP in reduced dimension slot

Remove tsne from Dataset SRP224648
```{r}
reducedDims(sce_list$SRP224648)$TSNE <- NULL
names(reducedDims(sce_list$SRP224648))
```
Now everything aligns and sce list can be collapsed into one sce object

```{r}
# batch identifier already present in coldata()$Dataset

# merge
sce <- do.call(cbind, sce_list)
# do.call expands list into individual elements for cbind to bind together

# remove sce_list to save memory
rm(sce_list)
```
The different datasets are now combined into one single cell experiment object. To remove technical variation between datasets, batch correction should be performed

## Pre-correction Visualization with TSNE plots

Before applying batch correction is performed, the data is visualized with TSNE plots. This way the effect of batch correction can be examined trough comparison

```{r}
# Feature selection for dimensionality reduction: want genes that are 
# informative of the biological structure of the data
dec <- modelGeneVar(sce, block = sce$Dataset) # decomposition of variance
# select genes with biological variance among datasets
chosen.hvgs <- getTopHVGs(dec, n = nrow(assay(sce))*0.1)
```

Function to plot variances of PCs to determine how many PCs to use for TSNE
```{r}
plot_pca_variances <- function(sce, n = 30, title = "Variance explained per PC") {

    # Extract percentage of variance explained
    var_exp <- attr(reducedDim(sce, "PCA"), "percentVar")

    # Prepare a data frame for ggplot
    df <- data.frame(
        PC = seq_along(var_exp),
        Variance = var_exp
    )
    # Limit the number of PCs plotted
    df <- df[seq_len(min(n, nrow(df))), ]
    
    ggplot(df, aes(x = PC, y = Variance)) +
        geom_bar(stat = "identity", fill = "steelblue") +
        geom_point() +
        theme_bw() +
        ylab("Explained variance (%)") +
        xlab("Principal component") +
        ggtitle(title)
}
```
```{r}
# using top 10% of all genes
if (FALSE) {
  sce <- runPCA(sce, subset_row = chosen.hvgs, ntop = nrow(assay(sce))*0.1)
  percent.var <- attr(reducedDim(sce, "PCA"), "percentVar")
  chosen.elbow <- findElbowPoint(percent.var)
  p <- plot_pca_variances(sce) + 
    geom_vline(xintercept = chosen.elbow, color = "red")
  ggsave(here("plots_and_figures", "PCA_explVar_uncorrected.png"), p)
}
```


```{r}
if (FALSE) {
  sce <- runTSNE(sce, dimred = "PCA", ntop = chosen.elbow)
  saveRDS(sce, file = "uncorrected_sce_RedDim.rds")
} else {
  sce <- readRDS("uncorrected_sce_RedDim.rds")
}

sce$Dataset = as.factor(sce$Dataset)
```


making dataframe for TSNE plots
```{r}
color_vec <- c("#5A5156", "#E4E1E3", "#F6222E", "#FE00FA", "#16FF32", "#3283FE", 
               "#FEAF16", "#B00068", "#1CFFCE", "#90AD1C", "#2ED9FF", "#DEA0FD", 
               "#AA0DFE", "#F8A19F", "#325A9B", "#C4451C", "#1C8356", "#B5EFB5", 
               "#B10DA1", "#3B00FB", "#1CBE4F", "#FA0087", "#333333", "#F7E1A0", 
               "#C075A6", "#782AB6", "#AAF400", "#BDCDFF", "#822E1C", "#85660D", 
               "#7ED7D1", "#1C7F93", "#D85FF7", "#683B79", "#66B0FF", "#FBE426")

TSNE_df <- as.data.frame(reducedDim(sce, "TSNE")) %>%  # make df from UMAP coordinates
  mutate(celltype = colData(sce)$Celltype) %>% # add celltype
  mutate(dataset = colData(sce)$Dataset) %>%   # add dataset id
  mutate(tissue = colData(sce)$Tissue)         # add tissue
```

General TSNE plot
```{r}
tsne_general_dataset <- TSNE_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )

tsne_general_celltype <- TSNE_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    scale_color_manual(values = color_vec) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                     # separate aes for legend dots
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
    theme(
      legend.text = element_text(size = 8),     # textsize of legend elements
      legend.key.size = unit(0.3, "cm"))# boxsize of legend elements
if (FALSE){
  outfile <- file.path(out_dir, paste0("TSNE_uncorrected_batch", ".png"))
  ggsave(outfile, tsne_general_dataset)
  outfile <- file.path(out_dir, paste0("TSNE_uncorrected_celltype", ".png"))
  ggsave(outfile, tsne_general_celltype)
}

c(tsne_general_dataset,tsne_general_celltype)
```
Possible problem: shared celltypes between datasets are taken together (eg xylem). Do we want this? Otherwise a suffix of the dataset should be added to all coldata values (eg xylem in colData$Celltype--> xylem_SRP272727_23_26, xylem_SRP145013, xylem_SRP335180)

Sample effects; most variance is due to differences between datasets, not celltypes
further verify this by plotting TSNE only for leaf datasets and for shared cell types 
```{r}
tsne_leaf_dataset <- TSNE_df %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) 
tsne_leaf_celltype <- TSNE_df %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )
if (FALSE){
  outfile <- file.path(out_dir, paste0("TSNE_uncorrected_batch_leaf", ".png"))
  ggsave(outfile, tsne_leaf_dataset)
  outfile <- file.path(out_dir, paste0("TSNE_uncorrected_celltype_leaf", ".png"))
  ggsave(outfile, tsne_leaf_celltype)
  }
c(tsne_leaf_dataset, tsne_leaf_celltype)
```
shared cell types between datasets dont cluster together --> strong dataset effect

same for the two root samples
```{r}
tsne_root_dataset <- TSNE_df %>%
  filter(str_detect(tissue, "Root")) %>%      # plot leaf datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) 
tsne_root_celltype <- TSNE_df %>%
  filter(str_detect(tissue, "Root")) %>%      # plot leaf datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )
if (FALSE){
  outfile <- file.path(out_dir, paste0("TSNE_uncorrected_batch_root", ".png"))
  ggsave(outfile, tsne_root_dataset)
  outfile <- file.path(out_dir, paste0("TSNE_uncorrected_celltype_root", ".png"))
  ggsave(outfile, tsne_root_celltype)
}

c(tsne_root_dataset, tsne_root_celltype)
```

Removing reduced dims to make some space
```{r}

reducedDim(sce, "PCA") <- NULL
reducedDim(sce, "TSNE") <- NULL
reducedDim(sce, "UMAP") <- NULL
sce_copy <- sce
```

## Batch correction

batch correction options:

- **linear modelling**,eg removeBatchEffect from limma, regressBatches from batchelor):
fits linear model trhough gene expression profile with batch term and then recomputing the observations without the batch effect
not good because assumes that composition of cell subpopulations is the same across batches
However, it is possible to take up cell type as factor in linear modelling (design parameter) --> could be good option

- **rescaleBatches** from batchelor
roughly equivalent to applying linear regression, also assumes constant composition -> not good

- **MNN correction**, eg fastMNN from batchelor
Finds nearest neighbor cell pairs between different batches --> assumes these cells are from the same biological state. Batch effect is estimated by the difference between these nearest neighbors. Does not assume that the population composition is the same or known beforehand. Key assumption is that the batch effect is orthogonal to the biology, usually reasonable as random vector is likely orthogonal --> could also be good option, but I am not sure if its a problem for this algorithm that not all celltypes are present in all datasets


## Batch correction with linear modelling - 

One think to keep in mind with this method, is that if a celltypes occurs only in a few batches or/and in low number in those batches the correction will be unstable. 
Therefore

Visualisation of overlap
```{r}
#table that counts the datasets containing a certain celltype 
cell_types_df <- as.data.frame(cell_types) |> 
  pivot_longer(cols = everything(), 
               names_to = 'dataset',
               values_to = 'celltype') |>
  group_by(celltype) |>
  summarise('#datasets containing celltype' = length(unique(dataset)), 
            datasets = paste(unique(dataset), collapse = ", ")) |>
  ungroup() |>
  filter(!celltype %in% c("/", "Unknown", "Unknow", "Contamination"))
#saveRDS(cell_types_df, file = file.path(out_dir, "_overlap.rds"))
cell_types_df
```

QuickCorrect function performs data set up steps:
- subsets all batches to common "universe" of features (dont really understand what this means)
- rescaling and within dataset normalisation to get logcounts
- feature selection to get hvgs
and then performs batch correction according to specified PARAM
```{r}
set.seed(73928)

correct_lm <- quickCorrect(
  sce, 
  batch = sce$Dataset,
  assay.type = "counts",   # expects counts assay not logcounts
  PARAM = RegressParam(    # use regressBatches as batch correction
    design = model.matrix(~ 0 + sce$Celltype) # no intercept
  )
)

assay(sce, "corrected_lm") <- assay(correct_lm$corrected) 
rm(correct_lm)
```
?maybe do visualisation on correct_lm$dec?


Visualisation with TSNEs of batch corrected data


```{r}
# using top 10% of all genes
if (FALSE) {
  sce <- runPCA(sce,
                exprs_values = "corrected_lm",
                subset_row = chosen.hvgs)
  
  percent.var <- attr(reducedDim(sce, "PCA"), "percentVar")
  chosen.elbow <- findElbowPoint(percent.var)
  
  p <- plot_pca_variances(sce) + 
    geom_vline(xintercept = chosen.elbow, color = "red")
  ggsave(here("plots_and_figures", "PCA_explVar_corrected_lm.png"), p)
}
```

```{r}
# saving because runTSNE takes long
if (FALSE) {
  sce <- runTSNE(sce, dimred = "PCA", ntop = chosen.elbow)
  saveRDS(sce, file = "corrected_lm_RedDim.rds")
} else {
  sce <- readRDS("corrected_lm_RedDim.rds")
}
```

```{r}
TSNE_df2 <- data.frame(
  TSNE1 = reducedDims(sce)$TSNE[,1],
  TSNE2 = reducedDims(sce)$TSNE[,2],
  celltype = sce$Celltype, 
  dataset = sce$Dataset,
  tissue = sce$Tissue
)
```

```{r}
if (FALSE) {
  saveRDS(TSNE_df2, file = "TSNE_df2.rds")
} else {
  TSNE_df2 <- readRDS("TSNE_df2.rds")
}
```

General TSNE plot
```{r}
tsne_batchcorlm_dataset <- TSNE_df2 %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )

tsne_batchcorlm_celltype <- TSNE_df2 %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    scale_color_manual(values = color_vec) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                     # separate aes for legend dots
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
    theme(
      legend.text = element_text(size = 8),     # textsize of legend elements
      legend.key.size = unit(0.3, "cm")) # boxsize of legend elements
outfile <- file.path(out_dir, paste0("tsne_batchcorlm_dataset", ".png"))
ggsave(outfile, tsne_batchcorlm_dataset)
outfile <- file.path(out_dir, paste0("tsne_batchcorlm_celltype", ".png"))
ggsave(outfile, tsne_batchcorlm_celltype)
c(tsne_batchcorlm_dataset, tsne_batchcorlm_celltype)
```
borders between dataset clusters a bit blurred, but still very apparent dataset effect. Little clustering on celltype

```{r}
tsne_batchcorlm_leaf_dataset <- TSNE_df2 %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) 
tsne_batchcorlm_leaf_celltype <- TSNE_df2 %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )
outfile <- file.path(out_dir, paste0("tsne_batchcorlm_leaf_dataset", ".png"))
ggsave(outfile, tsne_batchcorlm_leaf_dataset)
outfile <- file.path(out_dir, paste0("tsne_batchcorlm_leaf_celltype", ".png"))
ggsave(outfile, tsne_batchcorlm_leaf_celltype)
c(tsne_batchcorlm_leaf_dataset, tsne_batchcorlm_leaf_celltype)
```
looks a little better, mesophyll clusters together a bit. For the rest still bad


```{r}
tsne_batchcorlm_root_dataset <- TSNE_df2 %>%
  filter(str_detect(tissue, "Root")) %>%      # plot root datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) 

tsne_batchcorlm_root_celltype <- TSNE_df2 %>%
  filter(str_detect(tissue, "Root")) %>%      # plot root datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )
outfile <- file.path(out_dir, paste0("tsne_batchcorlm_root_dataset", ".png"))
ggsave(outfile, tsne_batchcorlm_root_dataset)
outfile <- file.path(out_dir, paste0("tsne_batchcorlm_root_celltype", ".png"))
ggsave(outfile, tsne_batchcorlm_root_celltype)
c(tsne_batchcorlm_root_dataset, tsne_batchcorlm_root_celltype)
```
Does not look any better


## Batch correction with MNN
start from uncorrected sce again, or work with counts in corrected sce 
- Compared to linear regression, MNN correction does not assume that the population composition is the same or known beforehand. This is because it learns the shared population structure via identification of MNN pairs and uses this information to obtain an appropriate estimate of the batch effect.
- Consider a cell a in batch A, and identify the cells in batch B that are nearest neighbors to a in the expression space defined by the selected features. Repeat this for a cell b in batch B, identifying its nearest neighbors in A. Mutual nearest neighbors are pairs of cells from different batches that belong in each other’s set of nearest neighbors. The reasoning is that MNN pairs represent cells from the same biological state prior to the application of a batch effect 
- key assumption: the batch effect is orthogonal to the biology in high-dimensional expression space.
- The most relevant parameter for tuning fastMNN() is k, which specifies the number of nearest neighbors to consider when defining MNN pairs. This can be interpreted as the minimum anticipated frequency of any shared cell type or state in each batch. Increasing k will generally result in more aggressive merging as the algorithm is more generous in matching subpopulations across batches. It can occasionally be desirable to increase k if one clearly sees that the same cell types are not being adequately merged across batches.
- In the specific case of MNN correction, we can also set auto.merge=TRUE to allow it to choose the “best” order in which to perform the merges. This is slower but can occasionally be useful when the batches involved have very different cell type compositions.

the fastMNN() function performs PCA to reduce the dimensions beforehand and speed up the downstream neighbor detection steps.

```{r}
set.seed(1000101001)
#read in file, can take long
if (FALSE) {mnn.out <- fastMNN(assay(sce_copy, 'counts'), 
                               batch = sce_copy$Dataset, d=50, k=20, subset.row=chosen.hvgs,
                               BSPARAM=BiocSingular::RandomParam(deferred=TRUE), 
                               auto.merge = TRUE)
  saveRDS(mnn.out, file = "corrected_mnn.rds")
} else {mnn.out <- readRDS("corrected_mnn.rds")}
#hier wil je de assay uit mnn.out in sce_copy krijgen om daar dan de TSNE op te laten lopen
#assay(sce_copy, "reconstructed") <- assay(mnn.out$reconstructed) 
```
--> not possible to put the assay of mnn.out in sce_copy because the dimensions are not the same and when making a new singleCellExperiment, you get 57.3GB, so not possible 

```{r}
#deze is niet nodig 
#set.seed(0010101010)
#if (TRUE) {
 # mnn.out <- runTSNE(mnn.out, dimred="corrected")
  #mnn.out$batch <- factor(mnn.out$batch)
  #saveRDS(mnn.out, file = "corrected_mnn_RedDim.rds")
#} else {
 # mnn.out <- readRDS("corrected_mnn_RedDim.rds")
#}
```

plot to check that celltypes cluster together   
```{r}
reducedDim(sce_copy, "corrected") <- reducedDim(mnn.out, "corrected")
set.seed(123)
sce_copy <- runTSNE(sce_copy, dimred = "corrected")
tsne_batchcormnn_celltype <- plotTSNE(sce_copy, colour_by = "Celltype") +
  scale_color_manual(values = color_vec) +
  labs(colour = 'Celltype')
tsne_batchcormnn_dataset <- plotTSNE(sce_copy, colour_by = "Dataset") +
  scale_color_manual(values = color_vec) +
  labs(colour = 'Dataset')
tsne_batchcormnn_celltype
tsne_batchcormnn_dataset
if (FALSE){outfile <- file.path(out_dir, paste0("tsne_batchcormnn_celltype", ".png"))
ggsave(outfile, tsne_batchcormnn_celltype)
outfile <- file.path(out_dir, paste0("tsne_batchcormnn_dataset", ".png"))
ggsave(outfile, tsne_batchcormnn_dataset)}
```
```{r}
# subset voor leaf
leaf_categories <- c("Leaf")
sce_leaf <- sce_copy[, sce_copy$Tissue %in% leaf_categories]
table(sce_leaf$Tissue)

tsne_batchcormnn_leaf_celltype <- plotTSNE(sce_leaf, colour_by = "Celltype") +
  scale_color_manual(values = color_vec) +
  labs(colour = "Celltype") +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))
tsne_batchcormnn_leaf_dataset <- plotTSNE(sce_leaf, colour_by = "Dataset") +
  scale_color_manual(values = color_vec) +
  labs(colour = "Dataset") +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))
tsne_leaf_celltype
tsne_leaf_dataset
if (FALSE) {
  outfile1 <- file.path(out_dir, "tsne_batchcormnn_leaf_celltype.png")
  ggsave(outfile1, tsne_batchcormnn_leaf_celltype, width = 8, height = 6, dpi = 300)

  outfile2 <- file.path(out_dir, "tsne_batchcormnn_leaf_dataset.png")
  ggsave(outfile2, tsne_batchcormnn_leaf_dataset, width = 8, height = 6, dpi = 300)
}

```
```{r}
# subset Root
sce_root <- sce_copy[, sce_copy$Tissue == "Root tip"]
table(sce_root$Tissue)
tsne_batchcormnn_root_celltype <- plotTSNE(sce_root, colour_by = "Celltype") +
  scale_color_manual(values = color_vec) +
  labs(colour = "Celltype") +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))
tsne_batchcormnn_root_dataset <- plotTSNE(sce_root, colour_by = "Dataset") +
  scale_color_manual(values = c("red", "blue")) +  # hier de kleuren instellen
  labs(colour = "Dataset") +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))
tsne_batchcormnn_root_celltype
tsne_batchcormnn_root_dataset
if (TRUE) {
  outfile1 <- file.path(out_dir, "tsne_batchcormnn_root_celltype.png")
  ggsave(outfile1, tsne_batchcormnn_root_celltype, width = 8, height = 6, dpi = 300)

  outfile2 <- file.path(out_dir, "tsne_batchcormnn_root_dataset.png")
  ggsave(outfile2, tsne_batchcormnn_root_dataset, width = 8, height = 6, dpi = 300)
}
```


TO DO: 

-Redo analysis with normalized counts from quickcluster
-Redo all TSNEs, only using the best PCs from PCA, also redo PCAs for it with 
top 10% of hvg instead of top 1000
  - batch correction lm method
  - batch corretion MNN method

- make matrix aggregating counts per celltype by mean, median, sum
- investigate what n_experiments is (see line 56), if these are multiple samples within each dataset we might have to correct for batch effects within each dataset aswell?


- additional data visualization plots?
- within dataset quality control plots (doublets, low quality droplets); see tutorial or stat genomics
- tidy code and comments


QUESTIONS:



```{r}
sessionInfo()
```




