---
title: "Chapter 1: Expression Data Acquisition and Preprocessing"
authors: 
"Eli Galle,
Emmeline Van Rijswijck,
Noor Ottevaere,
Woowon Jang"
date: "2025-11-15"
output: html_document
---

In this chapter we will download and preprocess single cell expression data retrieved from the Single Cell Plant DataBase (scPlandDB).

```{r, message=FALSE}
library(zellkonverter)
library(SingleCellExperiment)
library(R.utils)
library(here)
library(tidyverse)
library(scales)
library(reticulate)
library(scuttle)
library(scater)
library(scran)
library(batchelor)
library(DropletUtils)
library(Rtsne)
library(DelayedArray)
library(ragg) #to save plots 
library(PCAtools)
library(flextable)
library(stringr)
library(igraph)
```

```{r}
# Make sure we can render plots on hpc
options(bitmapType = "cairo")
```

## 1.1 Datasets Exploration

Setting root directory to files
```{r}
data_dir <- here("scRNAseq_data", "zea_mays")
out_dir <- here("plots_and_figures")
```

Getting file directories
```{r}
files <- list.files(path=data_dir, pattern = "\\.rds$", full.names = TRUE)
```

Reading in RDS files into list of single cell experiments
```{r}
sce_list <- lapply(files, readRDS)
names(sce_list) <- sub("\\.rds$", "", basename(files))
```

Gathering info on the datasets
```{r}
sce_info <- sapply(sce_list, function(x){
  list(
    organ = unique(colData(x)$Organ),
    tissue =  unique(colData(x)$Tissue),
    age = unique(colData(x)$ACE),
    n_experiments = length(unique(colData(x)$Orig.ident)),  # not really sure what these are (technical replicates, biological replicates, ...?)
    n_cells = ncol(x),
    n_celltypes = length(unique(colData(x)$Celltype)),
    total_genes = nrow(x),
    Avg_genes = round(mean(colData(x)$nFeature_RNA)),
    Avg_library_size = round(mean(colData(x)$nCount_RNA))
  )
})

sce_info["tissue", "SRP325657_2"] <- "Leaf abaxial mature peel" # different suffixes here were different samples (all of leaf abaxial mature peel tissues)

sce_info <- apply(sce_info, c(1,2), function(x) as.character(x)) 
sce_info <- as.data.frame(sce_info)
print(sce_info)
```

Visualization
```{r}
# transfer into tidy format for ggplot visualization
sce_info$stat <- rownames(sce_info)
sce_info_tidy <- sce_info |>
  pivot_longer(cols = -stat, names_to = "sample", values_to = "value")
```

```{r}
tissue_map <- sce_info_tidy |>
  filter(stat == "tissue") |>
  select(sample, tissue = value)
```

```{r}
sce_info_tidy2 <- sce_info_tidy |>
  left_join(tissue_map, by = "sample")
```

```{r}
sce_info_plot <- sce_info_tidy2 %>%
  filter(stat %in% c("n_cells", 
                     "n_celltypes", 
                     "total_genes", 
                     "Avg_genes", 
                     "Avg_library_size", 
                     "n_experiments")) %>%
  ggplot(aes(x = sample, y = as.numeric(value))) +
  geom_col(fill = "steelblue") +
  facet_wrap(~stat, scales = "free_y") +
  scale_x_discrete(labels = sce_info_tidy2$tissue) +
  scale_y_continuous(breaks = pretty_breaks(n = 4)) +
  labs(x = NULL, y = NULL) +
  theme(
      plot.title = element_text(face = "bold", size = 24),
      axis.text.x = element_text(angle = 90, vjust = 0.3),
      panel.grid = element_blank(),
      axis.line = element_blank(),
      axis.ticks = element_blank(),
  ) 
sce_info_plot
```

Save the plots and figures in a folder 
```{r}
outfile <- file.path(out_dir, paste0("plot_", 'sce_info_',".png")) #to overwrite an other plot with the same name 
ggsave(filename = outfile, plot = sce_info_plot)
```

Check annotated celltypes in datasets
```{r}
cell_types <- sapply(sce_list, function(x){
  list(
    c(unique(colData(x)$Celltype), rep( "/", 13 - length(unique(colData(x)$Celltype)) ))
  )
})
cell_types <- as.data.frame(cell_types)
cell_types
```

UMAP plot of each dataset colored by cell type:
```{r}
for (i in 1:length(sce_list)) {
  plot <- plotReducedDim(sce_list[[i]], dimred = "UMAP", colour_by = "Celltype") +
    ggtitle(paste(names(sce_list[i])), sce_info["tissue", i])
  print(plot)
  #save plots
  if (FALSE) {
    sample_name <- names(sce_list[i])
    outfile <- file.path(out_dir, paste0("UMAP_", sample_name, ".png"))
    agg_png(outfile, width = 1600, height = 1200, res = 200)
    print(plot)  
    dev.off()
    }
} 
```
PCA plot of each dataset colored by cell type
```{r}
for (i in 1:length(sce_list)) {
  plot <- plotReducedDim(sce_list[[i]], dimred = "PCA", colour_by = "Celltype") +
    ggtitle(paste(names(sce_list[i])), sce_info["tissue", i])
  print(plot)
  #save plots
  if (FALSE) {
    sample_name <- names(sce_list[i])
    outfile <- file.path(out_dir, paste0("PCA_", sample_name, ".png"))
    agg_png(outfile, width = 1600, height = 1200, res = 200)
    print(plot)  
    dev.off()
    }
} 
```

Violin plot of number of expressed genes per dataset
```{r, fig.width=7, fig.height=8}
df <- map_df(names(sce_list), function(name) {
  sce <- sce_list[[name]]
  expr <- assay(sce, "counts") 

  tibble(
    dataset = name,
    n_genes = colSums(expr > 0),
    tissue = colData(sce)$Tissue 
  )
})

violin <- ggplot(df, aes(x = dataset, y = n_genes, fill = tissue)) +
  geom_violin(trim = FALSE) +
  scale_fill_manual(values= palette.colors()) +
  scale_y_log10() +
  coord_flip() +
  labs(y = "Number of expressed genes") +
  theme_bw() +
  ggtitle("Distribution of number of expressed genes for each dataset") +
  theme(
    panel.grid = element_blank(),
  )
plot(violin)

if (FALSE){
  outfile <- file.path(out_dir, paste0("Violinplot",".png"))
  ggsave(outfile, violin)
  }
```
We see that for example the leaf datasets are comparable in expressed genes even though they are two different datasets


## 1.2 Quality control 

### 1.2.1 Filtering of unwanted cells 

Total number of cells before filtering
```{r}
sum(as.numeric(sce_info["n_cells",-ncol(sce_info)]))
```
Regard the present genotypes
```{r}
print("Genotypes")
for (i in 1:9) {
  cat(names(sce_list)[i], ":", 
      paste(unique(colData(sce_list[[i]])$Genotype), collapse = ", "), 
      "\n", 
      sep =" ")
}
```
In SRP417893 two genotypes are present: B73 and B73 (ns); We will only use B73 (wild type)

Look at present conditions
```{r}
print("Conditions")
for (i in 1:9) {
  cat(names(sce_list)[i], ":", 
      paste(unique(colData(sce_list[[i]])$Condition), collapse = ", "), 
      "\n", 
      sep =" ")
}
```
In SRP266308_1 two conditions are present: Kn1-O/+  and normal cells; We will only use normal cells

In SRP335180 two conditions are present: nitrate+ and nitrate- plants 
(grown in medium with or without nitrate); only use nitrate+ since this is the normal condition.

Remove the unwanted cells
```{r}
bad_celltypes <- c("Unknow", 
                   "Contamination")

bad_genotypes <- c("B73 (ns)") 

bad_conditions <- c("Kn1-O/+",  
                    "Nitrate-")
sce_list <- lapply(sce_list, function(sce) {
  keep <- !colData(sce)$Celltype %in% bad_celltypes &
          !colData(sce)$Genotype %in% bad_genotypes &
          !colData(sce)$Condition %in% bad_conditions
  sce[, keep]
})

rm(bad_celltypes, bad_genotypes, bad_conditions)
```

```{r}
sum(as.numeric(lapply(sce_list, ncol)))
```
228359 cells remain after filtering unwanted cells.

### 1.2.2 Filtering of low quality cells 

Check for empty droplets
```{r}
plots <- lapply(seq_along(sce_list), function(i) {
    sce <- sce_list[[i]]
    # Compute barcode ranks
    bcrank <- barcodeRanks(assay(sce, "counts"))

    uniq <- !duplicated(bcrank$rank)

    p <- ggplot(bcrank[uniq, ], aes(rank, total)) +
        geom_point() +
        scale_x_log10() +
        scale_y_log10() +
        labs(
            title = paste("Barcode Ranks:", names(sce_list)[i]),
            y = "Total UMI count"
        ) +
       theme(
            panel.background = element_blank(),   # remove panel background
            panel.grid.major = element_blank(),   # remove major grid lines
            panel.grid.minor = element_blank(),   # remove minor grid lines
            axis.line = element_line(color = "black") # add axis lines
        )
    p
})

# Print and save all plots
for (i in seq_along(plots)) {
  p <- plots[[i]]
  print(p)
  if (FALSE) {
    sample_name <- names(sce_list)[i]
    outfile <- file.path(out_dir, paste0("Barcode Ranks_", sample_name, ".png"))
    ggsave(outfile, p)
    }
}

rm(plots, p)
```
Filtering of empty droplets is already done
Mitochondrial DNA is also already filtered out (see paper scPlantDB)

### 1.2.3 Normalisation and log transformation

Check if counts are integers or log transformed
```{r}
check_normalization_sparse <- function(sce) {
  
  vals <- assay(sce, "counts")
  is_integer <- all(vals@x == floor(vals@x)) 
  
  # check for size factors
  sf_present <- !is.null(sizeFactors(sce))
  # output
  list(
    is_integer = is_integer,
    size_factors_present = sf_present
  )
}

# Apply to all SCEs in sce_list
norm_status <- lapply(names(sce_list), function(nm) {
  status <- check_normalization_sparse(sce_list[[nm]])
})

# Convert to data frame
norm_df <- do.call(rbind, lapply(norm_status, function(x) as.data.frame(t(x), stringsAsFactors=FALSE)))
norm_df_unlisted <- as.data.frame(apply(norm_df, c(1, 2), FUN = function(x) as.character(x)))
norm_df_unlisted

rm(norm_df, norm_df_unlisted)
```
All datasets contain integers and don't have size factors present. 
This is an indication that the data was not yet normalised.

Check if normalisation has already been done by plotting mean-variance relationship
```{r}
for (name in names(sce_list)) {
  sce <- sce_list[[name]]
  
  mat <- DelayedArray(assay(sce, "counts"))
  gene_means <- rowMeans(mat)
  gene_vars  <- rowVars(mat)
  
  # Free memory immediately after use
  rm(mat)
  gc()
  
  df <- data.frame(mean = gene_means, variance = gene_vars)
  
  p <- ggplot(df, aes(x = mean, y = variance)) +
    geom_point(alpha = 0.6, size = 0.7) +
    ggtitle(paste("Mean–Variance relationship:", name)) +
    scale_y_log10() + 
    scale_x_log10() +
    theme_bw() + 
    theme(
      panel.grid = element_blank()
    )
  
  print(p)
  if (TRUE) {
    sample_name <- name
    outfile <- file.path(out_dir, paste0("mean_var_before_normalisation_", sample_name, ".png"))
    ggsave(outfile, p)
    }
  # remove df aswell
  rm(df, gene_means, gene_vars, p)
}
rm(sce)
```
Linear relation between the variance in the mean: data is not logtransformed + normalised yet. This is done ourselves in the next section 

## 1.3 Within Dataset Normalisation 

First compute clusters using the quickcluster fucntion. Based on these clusters sizefactors are computed using the sizeFactors function.

!This may take some time. If you have already run the normalisation and saved the corresponding object, go to section 1.4.
```{r}
set.seed(42)
sce_list <- lapply(seq_along(sce_list), function(i) {
  sce <- sce_list[[i]]
  cl <- quickCluster(sce, assay.type = "counts")
  sce <- computeSumFactors(sce, clusters = cl, assay.type = "counts")
  
  sample_name <- names(sce_list)[i]
  outfile <- file.path(out_dir, paste0("deconvoluted_sizefactors_", sample_name, ".png"))
  
  png(outfile, width = 800, height = 600)
  p <- plot(
    librarySizeFactors(sce, assay.type = "counts"),
    sizeFactors(sce),
    xlab = "Library size factor",
    ylab = "Deconvolution size factor",
    log = "xy",
    pch = 16,
    col = as.integer(factor(cl))
  )
  abline(a = 0, b = 1, col = "red")
  
  dev.off()
  
  sce
})

```

Size factors are used to log transform and normalise the counts assay.
```{r}
sce_list <- lapply(sce_list, function(sce) {
logNormCounts(sce, assay.type = "counts")
})
```

```{r}
saveRDS(sce_list, file = "normalised_sce_list_quickcluster.rds")
```

Plot the mean-vor relation again.
```{r}
for (name in names(sce_list)) {
  sce <- sce_list[[name]]
  dec <- modelGeneVar(sce)
  varfit <- metadata(dec)
  
  sample_name <- name
  outfile <- file.path(out_dir, paste0("mean_var_after_normalisation_", sample_name, ".png"))
  
  png(outfile, width = 800, height = 600)
  plot(varfit$mean, varfit$var, 
      xlab="Mean of log-expression",
      ylab="Variance of log-expression")
  curve(varfit$trend(x), col="dodgerblue", add=TRUE, lwd=2)
  
  dev.off()
  rm(sce, dec, varfit)
}
```


## 1.4 Dataset Aggregation 

Read in the normalised singlecellExperiments.
```{r}
sce_list <- readRDS("normalised_sce_list_quickcluster.rds")
```

Here we will combine the seperate SingleCellExperiments into 1 object

First we verify if single cell experiments can be merged without problems, by investigating the metadatatable of the sce objects.
```{r}
# Verify if gene names align
print("gene names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(rownames(x), rownames(sce_list$SRP145013))))

# Verify if names in each slot of sce align
  # assays
print("assays names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(assays(x)), names(assays(sce_list$SRP145013)))))
  # colData
print("colData names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(assays(x)), names(assays(sce_list$SRP145013)))))
  # rowData
print("rowData names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(rowData(x)), names(rowData(sce_list$SRP145013)))))
  # reducedDims
print("reducedDims names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(reducedDims(x)), names(reducedDims(sce_list$SRP145013)))))
  # int_colData
print("int_colData names align:")
all(sapply(X = sce_list, 
       FUN = function(x) identical(names(int_colData(x)), names(int_colData(sce_list$SRP145013)))))
```

reduced dimensions do not align for all datasets
```{r}
sapply(X = sce_list, 
       FUN = function(x) names(reducedDims(x)))
```
Dataset SRP224648 has TSNE in addition to PCA and UMAP in reduced dimension slot

Remove tsne from Dataset SRP224648
```{r}
reducedDims(sce_list$SRP224648)$TSNE <- NULL
names(reducedDims(sce_list$SRP224648))
```
Now everything aligns and sce list can be collapsed into one sce object.
```{r}
sce <- do.call(cbind, sce_list)

# remove sce_list to save memory
rm(sce_list)
```
The different datasets are now combined into one single cell experiment object. To remove technical variation between datasets, batch correction should be performed

### 1.4.1 Visualisation with TSNE plots

Pre-correction Visualization with TSNE plots

Before batch correction is performed, the data is visualized with TSNE plots. This way the effect of batch correction can be examined trough comparison of the TSNE plots before and after batch correction.
```{r}
# Feature selection for dimensionality reduction: want genes that are 
# informative of the biological structure of the data
dec <- modelGeneVar(sce, block = sce$Dataset) # decomposition of variance
# select genes with biological variance among datasets
chosen.hvgs <- getTopHVGs(dec, n = nrow(assay(sce))*0.1)
```

Function to plot variances of PCs to determine how many PCs to use for TSNE
```{r}
plot_pca_variances <- function(sce, n = 30, title = "Variance explained per PC") {

    # Extract percentage of variance explained
    var_exp <- attr(reducedDim(sce, "PCA"), "percentVar")

    # Prepare a data frame for ggplot
    df <- data.frame(
        PC = seq_along(var_exp),
        Variance = var_exp
    )
    # Limit the number of PCs plotted
    df <- df[seq_len(min(n, nrow(df))), ]
    
    ggplot(df, aes(x = PC, y = Variance)) +
        geom_bar(stat = "identity", fill = "steelblue") +
        geom_point() +
        theme_bw() +
        ylab("Explained variance (%)") +
        xlab("Principal component") +
        ggtitle(title)
}
```

run PCA using top 10% of all genes.

! This may take a wile! If you already ran PCA once skip to the next cell
```{r}
sce <- runPCA(sce, subset_row = chosen.hvgs)
percent.var <- attr(reducedDim(sce, "PCA"), "percentVar")
chosen.elbow <- findElbowPoint(percent.var)
p <- plot_pca_variances(sce) + 
  geom_vline(xintercept = chosen.elbow, color = "red")
ggsave(here("plots_and_figures", "PCA_explVar_uncorrected.png"), p)

```


```{r}
if (FALSE) {
  sce <- runTSNE(sce, dimred = "PCA", ntop = chosen.elbow)
  saveRDS(sce, file = "uncorrected_sce_RedDim.rds")
} else {
  sce <- readRDS("uncorrected_sce_RedDim.rds")
}

sce$Dataset = as.factor(sce$Dataset)
```

Make dataframe for TSNE plots
```{r}
color_vec <- c("#5A5156", "#E4E1E3", "#F6222E", "#FE00FA", "#16FF32", "#3283FE", 
               "#FEAF16", "#B00068", "#1CFFCE", "#90AD1C", "#2ED9FF", "#DEA0FD", 
               "#AA0DFE", "#F8A19F", "#325A9B", "#C4451C", "#1C8356", "#B5EFB5", 
               "#B10DA1", "#3B00FB", "#1CBE4F", "#FA0087", "#333333", "#F7E1A0", 
               "#C075A6", "#782AB6", "#AAF400", "#BDCDFF", "#822E1C", "#85660D", 
               "#7ED7D1", "#1C7F93", "#D85FF7", "#683B79", "#66B0FF", "#FBE426")

TSNE_df <- as.data.frame(reducedDim(sce, "TSNE")) %>%  # make df from UMAP coordinates
  mutate(celltype = colData(sce)$Celltype) %>% # add celltype
  mutate(dataset = colData(sce)$Dataset) %>%   # add dataset id
  mutate(tissue = colData(sce)$Tissue)         # add tissue
```

General TSNE plot
```{r}
tsne_general_dataset <- TSNE_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )

tsne_general_celltype <- TSNE_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    scale_color_manual(values = color_vec) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                     # separate aes for legend dots
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
    theme(
      legend.text = element_text(size = 8),     # textsize of legend elements
      legend.key.size = unit(0.3, "cm"))# boxsize of legend elements

outfile <- file.path(out_dir, paste0("TSNE_uncorrected_batch", ".png"))
ggsave(outfile, tsne_general_dataset, width = 8, height = 6, units = "in")
outfile <- file.path(out_dir, paste0("TSNE_uncorrected_celltype", ".png"))
ggsave(outfile, tsne_general_celltype, width = 10, height = 6, units = "in")


c(tsne_general_dataset,tsne_general_celltype)
```
Sample effects; most variance is due to differences between datasets, not celltypes
further verify this by plotting TSNE only for leaf datasets and for shared cell types 
```{r}
tsne_leaf_dataset <- TSNE_df %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) 
tsne_leaf_celltype <- TSNE_df %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )

outfile <- file.path(out_dir, paste0("TSNE_uncorrected_batch_leaf", ".png"))
ggsave(outfile, tsne_leaf_dataset, width = 8, height = 6, units = "in")
outfile <- file.path(out_dir, paste0("TSNE_uncorrected_celltype_leaf", ".png"))
ggsave(outfile, tsne_leaf_celltype, width = 8, height = 6, units = "in")

c(tsne_leaf_dataset, tsne_leaf_celltype)
```
We observe a large effect due to dataset, not celltype.

Verify this affect for the root datasets too.
```{r}
tsne_root_dataset <- TSNE_df %>%
  filter(str_detect(tissue, "Root")) %>%      # plot leaf datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) 
tsne_root_celltype <- TSNE_df %>%
  filter(str_detect(tissue, "Root")) %>%      # plot leaf datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    )

outfile <- file.path(out_dir, paste0("TSNE_uncorrected_batch_root", ".png"))
ggsave(outfile, tsne_root_dataset, width = 8, height = 6, units = "in")
outfile <- file.path(out_dir, paste0("TSNE_uncorrected_celltype_root", ".png"))
ggsave(outfile, tsne_root_celltype, width = 8, height = 6, units = "in")

c(tsne_root_dataset, tsne_root_celltype)
```

Removing reduced dims to free memory.
```{r}
reducedDim(sce, "PCA") <- NULL
reducedDim(sce, "TSNE") <- NULL
reducedDim(sce, "UMAP") <- NULL
sce_copy <- sce
```

## 1.6 Batch correction

Possible batch correction methods:

- **linear modelling**,eg removeBatchEffect from limma, regressBatches from batchelor):
fits linear model through gene expression profile with batch term and then recomputing the observations without the batch effect.
Assumes that composition of cell subpopulations is the same across batches

- **rescaleBatches** from batchelor
roughly equivalent to applying linear regression, also assumes constant composition.

- **MNN correction**, eg fastMNN from batchelor
Finds nearest neighbor cell pairs between different batches. Assumes these cells are from the same biological state. Batch effect is estimated by the difference between these nearest neighbors. Does not assume that the population composition is the same or known beforehand. Key assumption is that the batch effect is orthogonal to the biology, usually reasonable as random vector is likely orthogonal --> could also be good option, but I am not sure if its a problem for this algorithm that not all celltypes are present in all datasets


### 1.6.1 Explore data to determine optimal batch correction method
Visualisation of overlap
```{r}
#table that counts the datasets containing a certain celltype 
cell_types_df <- as.data.frame(cell_types) |> 
  pivot_longer(cols = everything(), 
               names_to = 'dataset',
               values_to = 'celltype') |>
  group_by(celltype) |>
  summarise('#datasets containing celltype' = length(unique(dataset)), 
            datasets = paste(unique(dataset), collapse = ", ")) |>
  ungroup() |>
  filter(!celltype %in% c("/", "Unknown", "Unknow", "Contamination"))
#saveRDS(cell_types_df, file = file.path(out_dir, "_overlap.rds"))
cell_types_df
```
composition of celltypes among different datasets is defenitely not constant --> cannot use regressBatches or rescaleBatches

```{r}
cell_types_long <- as.data.frame(cell_types) |>
  pivot_longer(
    cols = everything(),
    names_to = "Dataset",
    values_to = "CellType"
  ) |>
  filter(!CellType %in% c("/", "Unknown", "Unknow", "Contamination"))

dataset_pairs <- expand_grid(
  Dataset_1 = unique(cell_types_long$Dataset),
  Dataset_2 = unique(cell_types_long$Dataset)
) |>
  filter(Dataset_1 < Dataset_2)  # avoid dubbel & zelf-comparisons

pairwise_overlap <- dataset_pairs |>
  rowwise() |>
  mutate(
    Shared_CellTypes = list(
      intersect(
        cell_types_long$CellType[cell_types_long$Dataset == Dataset_1],
        cell_types_long$CellType[cell_types_long$Dataset == Dataset_2]
      )
    ),
    `# Shared Cell Types` = length(Shared_CellTypes),
    Overlap = `# Shared Cell Types` > 0
  ) |>
  ungroup() |>
  mutate(
    `Shared Cell Types` = sapply(Shared_CellTypes, paste, collapse = ", ")
  ) |>
  select(
    Dataset_1,
    Dataset_2,
    `# Shared Cell Types`,
    `Shared Cell Types`,
    Overlap
  )

ft_pairs <- flextable(pairwise_overlap) |>
  autofit() |>
  theme_vanilla()

save_as_docx(ft_pairs, path = "pairwise_dataset_overlap.docx")
ft_pairs
```
```{r}
g <- graph_from_data_frame(
  pairwise_overlap |>
    filter(Overlap) |>
    select(Dataset_1, Dataset_2),
  directed = FALSE
)

cliques <- maximal.cliques(g)

dataset_overlap_sets <- tibble(
  Set_ID = seq_along(cliques),
  Datasets = sapply(
    cliques,
    function(x) paste(V(g)$name[x], collapse = ", ")
  ),
  `# Datasets in Set` = sapply(cliques, length)
) |>
  arrange(desc(`# Datasets in Set`))

ft_sets <- flextable(dataset_overlap_sets) |>
  autofit() |>
  theme_vanilla()

save_as_docx(ft_sets, path = "dataset_overlap_sets.docx")
ft_sets
```
```{r}
dataset_in_overlap_set <- tibble(
  Dataset = unique(c(pairwise_overlap$Dataset_1, pairwise_overlap$Dataset_2))
) |>
  mutate(
    In_Overlap_Set = Dataset %in%
      unique(unlist(lapply(cliques, function(x) V(g)$name[x])))
  ) |>
  arrange(desc(In_Overlap_Set), Dataset)

dataset_in_overlap_set
```
Transitive connectivity between datasets is sufficient, meaning that not every dataset pair is required to share a cell type directly, as long as all datasets are connected through a chain of overlapping cell types. The assumption of overlap between datasets is met, we can do MNN.

## 1.6.2 Batch correction with MNN

Compared to linear regression, MNN correction does not assume that the population composition is the same or known beforehand. This is because it learns the shared population structure via identification of MNN pairs and uses this information to obtain an appropriate estimate of the batch effect.

Consider a cell a in batch A, and identify the cells in batch B that are nearest neighbors to a in the expression space defined by the selected features. Repeat this for a cell b in batch B, identifying its nearest neighbors in A. Mutual nearest neighbors are pairs of cells from different batches that belong in each other’s set of nearest neighbors. The reasoning is that MNN pairs represent cells from the same biological state prior to the application of a batch effect 

Key assumption: the batch effect is orthogonal to the biology in high-dimensional expression space.

The most relevant parameter for tuning fastMNN() is k, which specifies the number of nearest neighbors to consider when defining MNN pairs. This can be interpreted as the minimum anticipated frequency of any shared cell type or state in each batch. Increasing k will generally result in more aggressive merging as the algorithm is more generous in matching subpopulations across batches. It can occasionally be desirable to increase k if one clearly sees that the same cell types are not being adequately merged across batches.

In the specific case of MNN correction, we can also set auto.merge=TRUE to allow it to choose the “best” order in which to perform the merges. This is slower but can occasionally be useful when the batches involved have very different cell type compositions.

the fastMNN() function performs PCA to reduce the dimensions beforehand and speed up the downstream neighbor detection steps.

This again may time some time, if you already ran the batchcorrection once, just read in the saved .rds file.
```{r}
set.seed(1000101001)
if (FALSE) {mnn.out <- fastMNN(assay(sce_copy, 'counts'), 
                               batch = sce_copy$Dataset, d=50, k=20, subset.row=chosen.hvgs,
                               BSPARAM=BiocSingular::RandomParam(deferred=TRUE), 
                               auto.merge = TRUE)
  saveRDS(mnn.out, file = "corrected_mnn.rds")
} else {
  mnn.out <- readRDS("corrected_mnn.rds")
}
```


```{r}
if (FALSE) {
  sce2 <- runPCA(mnn.out,
                exprs_values = "reconstructed",
                subset_row = chosen.hvgs)
  percent.var <- attr(reducedDim(sce2, "PCA"), "percentVar")
  chosen.elbow <- findElbowPoint(percent.var)
  saveRDS(sce2, file = "PCA_dimred_mnn.rds")
  p2 <- plot_pca_variances(sce2) + 
    geom_vline(xintercept = chosen.elbow, color = "red")
  ggsave(here("plots_and_figures", "PCA_explVar_corrected_mnn.png"), p2)
} else {
  sce2 <- readRDS("PCA_dimred_mnn.rds")
}
```

```{r}
if (FALSE) {
  sce2 <- runTSNE(sce2, dimred = "PCA", ntop = chosen.elbow)
  saveRDS(sce2, file = "corrected_mnn_RedDim.rds")
} else {
  sce2 <- readRDS("corrected_mnn_RedDim.rds")
}
```

```{r}
TSNE_df2_mnn <- data.frame(
  TSNE1 = reducedDims(sce2)$TSNE[,1],
  TSNE2 = reducedDims(sce2)$TSNE[,2],
  celltype = sce$Celltype, 
  dataset = sce$Dataset,
  tissue = sce$Tissue
)
```

plot to check that celltypes cluster together   
```{r}
tsne_batchcormnn_dataset <- TSNE_df2_mnn %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
  theme(legend.position = "none") +
  ggtitle("Coloured by dataset")

tsne_batchcormnn_celltype <- TSNE_df2_mnn %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    scale_color_manual(values = color_vec) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                     # separate aes for legend dots
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
    theme(
      legend.text = element_text(size = 8),     # textsize of legend elements
      legend.key.size = unit(0.3, "cm")) +
  theme(legend.position = "none") + 
  ggtitle("Coloured by celltype")

# Save plots
outfile <- file.path(out_dir, paste0("tsne_batchcormnn_dataset", ".png"))
ggsave(outfile, tsne_batchcormnn_dataset)
outfile <- file.path(out_dir, paste0("tsne_batchcormnn_celltype", ".png"))
ggsave(outfile, tsne_batchcormnn_celltype)
c(tsne_batchcormnn_dataset, tsne_batchcormnn_celltype)
```

```{r}
tsne_batchcormnn_leaf_dataset <- TSNE_df2_mnn %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
  theme(legend.position = "none") + 
  ggtitle("Coloured by dataset")

tsne_batchcormnn_leaf_celltype <- TSNE_df2_mnn %>%
  filter(str_detect(tissue, "Leaf")) %>%      # plot leaf datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
  theme(legend.position = "none") + 
  ggtitle("Coloured by celltype")

# Save plots
outfile <- file.path(out_dir, paste0("tsne_batchcormnn_leaf_dataset", ".png"))
ggsave(outfile, tsne_batchcormnn_leaf_dataset)
outfile <- file.path(out_dir, paste0("tsne_batchcormnn_leaf_celltype", ".png"))
ggsave(outfile, tsne_batchcormnn_leaf_celltype)
c(tsne_batchcormnn_leaf_dataset, tsne_batchcormnn_leaf_celltype)
```



```{r}
tsne_batchcormnn_root_dataset <- TSNE_df2_mnn %>%
  filter(str_detect(tissue, "Root")) %>%      # plot root datasets
  group_by(celltype) %>%                      
  filter(n_distinct(dataset) > 1) %>%
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = dataset)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
  theme(legend.position = "none") + 
  ggtitle("Coloured by dataset")

tsne_batchcormnn_root_celltype <- TSNE_df2_mnn %>%
  filter(str_detect(tissue, "Root")) %>%      # plot root datasets
  group_by(celltype) %>%
  filter(n_distinct(dataset) > 1) %>% 
  ungroup() %>%
  ggplot(aes(x = TSNE1, y = TSNE2, color = celltype)) +
    geom_point(size = 0.01, alpha = 0.1) +
    labs(x = "TSNE1", y = "TSNE2") +
    guides(                                   # separate aes for legend 
      color = guide_legend(override.aes = list(size=3, alpha = 1))
    ) +
  theme(legend.position = "none") + 
  ggtitle("Coloured by celltype")

# Save plots
outfile <- file.path(out_dir, paste0("tsne_batchcormnn_root_dataset", ".png"))
ggsave(outfile, tsne_batchcormnn_root_dataset)
outfile <- file.path(out_dir, paste0("tsne_batchcormnn_root_celltype", ".png"))
ggsave(outfile, tsne_batchcormnn_root_celltype)
c(tsne_batchcormnn_root_dataset, tsne_batchcormnn_root_celltype)
```

## 1.7 Data summary 

We will summarize the data across cell type and organ, enabling further analysis.

```{r}
sce <- readRDS("uncorrected_sce_RedDim.rds")
```

For later analysis, different cellphases will be grouped into 1 celltype
```{r}
# cell types annotated as being a certain cell phase will be grouped together
# original celltype is kept in Celltype_original

colData(sce)$Celltype_original <- colData(sce)$Celltype
sce$Celltype[sce$Celltype %in% c("S phase", "G2/M phase")] <- "Cell phase"
```

We don't need reduced dimensions for further analysis
```{r}
reducedDim(sce) <- NULL
```

Next we make a new colData variable indicating the cell type and organ combination for each cell. This is the variable across we will summarize our counts. 
```{r}
# string manipulation to make resulting variable more readable
# eg "RootCap_RootShoot" in stead of "Root cap_Root Shoot"

celltypes <- sce$Celltype %>%
  str_to_title() %>%
  str_remove_all(" ")

organs <- sce$Organ %>%
  str_to_title() %>%
  str_remove_all(" ")


# create new colData variable
sce$Celltype_Organ <- paste(celltypes, organs, sep="_")

rm(celltypes, organs)
```

Summarize/aggregate counts across Celltype_Tissue
```{r}
summarized_sce <- aggregateAcrossCells(
  sce, 
  ids = sce$Celltype_Organ,
  use.assay.type = c("counts","logcounts"),
  statistics = c("sum", "mean", "median"),
  store.number = "ncells")                   # store number of cells 
```

Save objects for further analysis as RDS
```{r}
saveRDS(sce, "uncorrected_sce.rds")
```
```{r}
saveRDS(summarized_sce, file = "summarized_sce.rds")
```

```{r}
sessionInfo()
```




