---
title: "chapter4"
output: html_document
execute: 
  echo: true
  eval: false
  warning: false
---

# 4 Expression divergence between duplicate pairs

In this chapter, we will explore expression divergence between duplicate
pairs using different metrics to investigate if there is biased divergence
depending on the duplication mechanisms. 

First load required packages:
```{r, warning=FALSE, message=FALSE}
library(exdiva)
library(SingleCellExperiment)
library(SuperCell)
library(SummarizedExperiment)
library(here)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(purrr)
library(HybridExpress)
library('ggh4x')
library('ggsci')
library(patchwork)
library('FSA')
```

```{r}
# Define color palette for duplication modes
dup_pal <- c(
    SD = "#000000",  
    TD = "#E69F00",
    PD = "#56B4E9",
    rTRD = "#009E73",
    dTRD = "#F0E442",
    DD = "#0072B2"
)
```

```{r}
# Plot background
bg <- grid::linearGradient(colorRampPalette(c("gray90", "white"))(100))
```

Read in the SingleCellExperiment created in chapter1

We need to do metacell aggregation to reduce noise in the data.
We do this using the SuperCell package.

# 4.1 Create the metacells

! You only have to run this once, if you already created the metacell_sce.rds file: go to 2

```{r}
sce <- readRDS(here("chapters", "sce_paralogs.rds"))
```

```{r}
sc <- SCimplify(
   X = assay(sce, "logcounts"),
   gamma = 30,
   cell.annotation = colData(sce)$Celltype_Organ
)
```

```{r}
length(sc$membership)
max(sc$membership)
```
We have aggregated 228 359 singlecells into 18 196 metacells

```{r}
metacell_logcounts <- supercell_GE(
  ge = assay(sce, "logcounts"),
  groups = sc$membership,
  mode = "sum"
)
  
metacell_sce <- SingleCellExperiment(
  assays = list(logcounts = metacell_logcounts)
)
colData(metacell_sce)$celltype <- sc$SC.cell.annotation. ## Add annotation celltype_organ to the metacells
```

```{r}
saveRDS(metacell_sce, here("chapters", "metacell_sce.rds"))
```

# 4.2 Comparing distributions of gene-gene correlations by duplication mode

read in the metacells and duplication data
```{r}
metacell_sce <- readRDS(here("chapters", "metacell_sce.rds"))
```

```{r}
paralog_pairs <- readRDS(here("Duplication_data", "duplication_info_filtered.rds"))

paralog_pairs$pairs %>%
     distinct(peak_name)
```
Filter out paralogue pairs with peak_name is NA
```{r}
paralog_pairs$pairs <- paralog_pairs$pairs[!is.na(paralog_pairs$pairs$peak_name), ]
```


```{r}
# Create a dataframe containing only the duplicated pairs and their type
dups <- as.data.frame(paralog_pairs$pairs[, c("dup1", "dup2", "type")])

# Calculate the Spearman's correlation coëfficiënt using the exdiva package
se_simple <- SummarizedExperiment(
  assays = list(logcounts = as.matrix(logcounts(metacell_sce)))
)

sp_coeff <- calculate_cor(
  se = se_simple,
  dups = dups,
  assay_name = "logcounts",
  method = "spearman"
)
```

Remove pairs where correlation is NA
```{r}
sp_coeff <- sp_coeff %>%
     filter(!is.na(cor))
```


Next, for each sample, we will test for significant differences by 
duplication mode using Kruskal-Wallis tests followed by a post-hoc Dunn's test. 
To account for differences in ages between gene pairs, we will do this 
separately for each age group.

```{r}
# Ad all info to the correlation dataframe
sp_coeff_age <- dplyr::inner_join(paralog_pairs$pairs, sp_coeff, by = c("dup1", "dup2", "type")) 
# Type can be added as an innerjoin key bc they will always be the same
# This way we will only have 1 type column in the result
```

Kruskal-Wallis test is a non-parametric statistical test to test for significant
differences between multiple groups.The test is based on the ranks of the data.
It is an extension of the Mann-Whitney-Wilcoxon test (only appropriate to compare 2 distr.)

Post-hoc Dunnn test is used after the Kruskal-Wallis test to calculate wether the 
pairwise differences in average rank is significant.

```{r}
# Credits to Fabricio Almeida Silva 
# Compare distros
#' Wrapper function to compare distros with Kruskal-Wallis + Dunn
cld_kw_dunn <- function(df, var = "type", value = "csum") {
    
    # Perform Kruskal-Wallis test followed by a post-hoc Dunn's test
    dunn <- FSA::dunnTest(
        df[[value]] ~ df[[var]], method = "bh"
    ) |>
        purrr::pluck("res") |>
        arrange(Z)
    
    # Get compact letter display (CLD)
    cld <- rcompanion::cldList(
        comparison = dunn$Comparison,
        p.value = dunn$P.adj,
        threshold = 0.05
    )
    
    return(cld)
}

compare_cors <- function(rho_df) {
    
    # Add a variable indicating sample + Ks peak
    rho_df$id <- paste0(rho_df$sample, "_", rho_df$peak)
    ids <- unique(rho_df$id)
    
    # Perform tests and get CLD for each sample-peak
    cld_df <- Reduce(rbind, lapply(ids, function(x) {
        
        fdf <- rho_df[rho_df$id == x, ]
        
        # Perform Kruskal-Wallis test followed by a post-hoc Dunn's test
        cld <- cld_kw_dunn(fdf, value = "cor")
        cld$sample <- unique(fdf$sample)
        cld$peak <- unique(fdf$peak)
        cld$peak_name <- unique(fdf$peak_name)
        
        return(cld)
    }))
    
    
    
    return(cld_df)
}
```

```{r, warning= FALSE, message=FALSE}
comp_cors <- compare_cors(sp_coeff_age) |> mutate(species = "Zma")
```

```{r}
# we add a dummy "sample" column so that the downstream compare_cors()
# function — which was written for multi-sample spatial data — works unchanged.
comp_cors$sample <- "Zma_metacell"
```

```{r}
# Plot heatmap with CLDs
p_heatmap_rho <- comp_cors |>
    mutate(Group = factor(Group, levels = rev(names(dup_pal)))) |>
    ggplot(aes(x = peak_name, y = Group, fill = Letter)) +
    geom_tile(color = "gray90", linewidth = 0.8, alpha = 0.7) +
    geom_text(aes(label = Letter), size = 3) +
    ggh4x::facet_nested(~ peak_name, scales = "free", space = "free") +
    theme_classic() +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    ) +
    labs(
        title = "Differences in gene-gene correlations by duplication mode",
        subtitle = "CLD from Kruskal-Wallis + post-hoc Dunn's test; P <0.05",
        x = "Peak", y = "Mode"
    ) +
    theme(
        legend.position = "none"
    ) +
    ggsci::scale_fill_npg()

p_heatmap_rho
```
POAC WGD: happened ∼53 MYA
ZEAM WGD: happened ∼16 MYA 

# 4.3 Fisher's z-transformed rho across the dataset

we compute the Fisher's z-transformed rho per pair and visualise the
distribution by duplication mode. 

Why we use this transformation:
When the Pearson correlation coëfficiënt is near -1/1 the distribution of the coëfficiënt is 
skewed. By applying the z-tranformation we compensate this affect and the 
distribution will become more normal.

```{r}
# Get mean Fisher's Z-transformed rho estimates

zrho <- sp_coeff_age |>
    mutate(
        cor = 0.5 * log((1 + cor) / (1 - cor)),  # Fisher's z-transformation
        pair = str_c(dup1, dup2, sep = "_"),
        species_peak = str_c("zma", peak, sep = "_")
    ) |>
    group_by(pair, peak) |>
    mutate(mean_cor = mean(cor)) |>
    ungroup() |>
    distinct(pair, peak, .keep_all = TRUE) |>
    mutate(
        type = factor(type, levels = names(dup_pal)),
        peak = as.character(peak)
    ) |>
    dplyr::select(-c(cor, pair))

# Compare distributions using the same KW + Dunn approach
zrho_cld <- lapply(
    split(zrho, zrho$species_peak),
    cld_kw_dunn, var = "type", value = "mean_cor"
) |>
    bind_rows(.id = "species_peak") |>
    tidyr::separate(species_peak, c("sp", "peak"), sep = "_") |>
    dplyr::rename(type = Group) |>
    mutate(
        type = factor(type, levels = names(dup_pal)),
        peak = as.character(peak)
    ) |>
    left_join(zrho |> dplyr::select(peak, peak_name) |> distinct())

# Plot distributions with CLDs
p_zrho <- zrho |>
    filter(mean_cor != Inf, !is.na(mean_cor)) |>
    mutate(peak_name = factor(peak_name, levels = sort(unique(peak_name)))) |>
    ggplot(aes(x = mean_cor, y = type)) +
    geom_violin(aes(fill = type), scale = "width", show.legend = FALSE) +
    scale_fill_manual(values = dup_pal) +
    geom_boxplot(width = 0.1, outlier.color = "gray60", outlier.alpha = 0.5) +
    geom_label(
        data = zrho_cld,
        aes(x = 0.8, y = type, label = Letter)
    ) +
    ggh4x::facet_nested(~ peak_name, scales = "free", space = "free") +
    labs(
        x = expression("Fisher's z-transformed" ~ rho), 
        y = NULL,
        title = "Gene-gene correlations and duplication mode",
        subtitle = "CLD from Kruskal-Wallis + post-hoc Dunn's test; P <0.05"
    ) +
    scale_x_continuous(limits = c(-0.5, 1)) +
    theme_classic()

p_zrho
```

Older duplicates consistently display higher expression divergence, in line with findings from previous studies. Duplication mode also plays a clear role in the recent duplicates: in maize, paralogs show the greatest expression divergence among recent duplicates, for example in SD, TD, and dTRD. This contrasts with previous work, where duplication modes that preserve cis-regulatory landscapes — such as WGD and TRD — were associated with lower expression divergence compared to modes that disrupt them.

# 4.4 Expression divergence over time

Here, we explore how gene-gene correlations change over time using $K_s$ as
a proxy for time. We use Fisher's z-transformed Spearman's $\rho$ and fit
both a linear (`lm(zrho ~ Ks)`) and a log-linear (`lm(zrho ~ log(Ks))`)
model to determine which better describes the relationship.
```{r}
# Get Fisher's Z-transformed rho for all duplicates, filtered to Ks < 3
mean_rho_zma <- sp_coeff_age |>
    filter(Ks < 3) |>
    mutate(
        cor = 0.5 * log((1 + cor) / (1 - cor)),
        pair = str_c(dup1, dup2, sep = "_")
    ) |>
    distinct(pair, .keep_all = TRUE) |>
    mutate(
        type = factor(type, levels = names(dup_pal))
    ) |>
    dplyr::select(zrho = cor, Ks, Ka, type) |>
    filter(!zrho %in% c(Inf, -Inf), !is.na(zrho))
```

```{r}
# Compare linear vs log-linear models per duplication mode
dmodes <- c("SD", "TD", "PD", "rTRD", "dTRD", "DD")

model_comp_zma <- Reduce(rbind, lapply(dmodes, function(y) {
    
    df_mode <- mean_rho_zma |>
        filter(type == y) |>
        mutate(logks = log(Ks))
    
    if (nrow(df_mode) < 5) return(NULL)  # skip modes with too few pairs
    
    m1 <- lm(zrho ~ Ks,    data = df_mode)
    m2 <- lm(zrho ~ logks, data = df_mode)
    
    data.frame(
        BIC1  = BIC(m1, m2)$BIC[1],
        BIC2  = BIC(m1, m2)$BIC[2],
        AIC1  = AIC(m1, m2)$AIC[1],
        AIC2  = AIC(m1, m2)$AIC[2],
        Rsq1  = summary(m1)$r.squared,
        Rsq2  = summary(m2)$r.squared,
        type  = y
    ) |>
        mutate(
            best_BIC = ifelse(BIC1 < BIC2, "linear", "log-linear"),
            best_AIC = ifelse(AIC1 < AIC2, "linear", "log-linear")
        )
}))

model_comp_zma
```

```{r}
# Scatterplot: z-rho ~ log(Ks) with linear trend per duplication mode
p_rho_ks_zma <- mean_rho_zma |>
    filter(!is.na(type)) |>
    mutate(logks = log(Ks)) |>
    ggplot(aes(x = logks, y = zrho)) +
    geom_point(aes(color = type), alpha = 0.4) +
    scale_color_manual(values = dup_pal) +
    geom_smooth(method = "lm", color = "firebrick3", se = TRUE) +
    ggpubr::stat_regline_equation(
        aes(label = ..eq.label..),
        label.x = -3.5, label.y = 2.5, size = 3
    ) +
    facet_wrap(~ type, nrow = 1) +
    coord_cartesian(ylim = c(-1, 3)) +
    scale_x_continuous(limits = c(-4, 1.1)) +
    theme_classic() +
    theme(legend.position = "none") +
    labs(
        title = "Expression divergence over time — Z. mays (single-cell metacells)",
        x = expression(log ~ K[s]),
        y = expression("Fisher's z-transformed" ~ rho)
    )

p_rho_ks_zma
```

## 4.5 Expression divergence by gene family

Here, we will explore how duplicates in each gene family diverge, looking
for families that tend to have more similar expression profiles (i.e., less
divergence), and families with more divergent expression profiles.
We will start by assigning families to each pair and ranking families by median
correlations.

```{r}
# Load gene family assignments from PLAZA
fams <- readRDS(here("chapters","gene_families.rds"))
fams <- fams |> filter(species == "zma")
```

```{r}
# Assign families to each pair
zrho_fams <- zrho |>
    inner_join(fams |> dplyr::select(dup1 = gene, family)) |>
    filter(!mean_cor %in% c(Inf, -Inf))
```
```{r}
# Get Q1, Q3, median, and rank families by median
fams_ranked <- zrho_fams |>
    group_by(family) |>
    summarise(
        q1 = quantile(mean_cor, 0.25, na.rm = TRUE),
        median = quantile(mean_cor, 0.5, na.rm = TRUE),
        q3 = quantile(mean_cor, 0.75, na.rm = TRUE)
    ) |>
    arrange(-median) |>
    mutate(family = factor(family, levels = family)) |>
    ungroup()
```

Next, we will visualize distributions of correlations for each family
using boxplots. We will use median correlations calculated above to sort
boxes from highest to lowest correlation.

```{r}
# Visualize IQRs of correlations per family
p_zrho_fam <- fams_ranked |>
    ggplot(aes(x = family)) +
    geom_linerange(
        aes(ymin = q1, ymax = q3),
        color = "gray20"
    ) +
    geom_point(aes(y = median), color = "firebrick", size = 0.5) +
    coord_cartesian(ylim = c(-0.2, 1.5)) +
    theme_classic() +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.background = element_rect(fill = bg)
    ) +
    labs(
        title = "Gene-gene correlations for each gene family",
        subtitle = "Families are ranked by median values (red dots)",
        x = "Ranked gene families",
        y = expression("IQR of Fisher's z-transformed" ~ rho)
    )

p_zrho_fam
```
```{r}
go_annot <- readRDS(here("Duplication_data", "functional_annotation.rds"))
```


```{r}
# Select top 10% of the families with highest median correlations
top10_high <- fams_ranked |>
    slice_max(median, prop = 0.1) |>
    ungroup()

top10_low <- fams_ranked |>
    slice_min(median, prop = 0.1) |>
    ungroup()

# Perform overrepresentation analyses
sp <- names(go_annot)
sea <- lapply(sp, function(x) {
    
    ## Get top 10% families (highest and lowest correlations)
    t10_high <- top10_high |>
        pull(family)
    
    t10_low <- top10_low |>
        pull(family)
    
    ## Get gene sets
    genes_high <- fams |>
        filter(family %in% t10_high) |>
        pull(gene)
    
    genes_low <- fams |>
        filter(family %in% t10_low) |>
        pull(gene)
    
    ## Perform SEA
    bg <- zrho_fams 
    bg <- unique(c(bg$dup1, bg$dup2))
    
    df_high <- HybridExpress::ora(
        genes = genes_high,
        annotation = as.data.frame(go_annot[[x]]),
        background = bg
    ) |>
        mutate(category = "highcor")
    
    df_low <- HybridExpress::ora(
        genes = genes_low,
        annotation = as.data.frame(go_annot[[x]]),
        background = bg
    ) |>
        mutate(category = "lowcor")
    
    final_df <- bind_rows(df_high, df_low)
    
    return(final_df)
}) |>
    bind_rows()
```

Logically the high correlation genes should be the most important and
fundamental functions. Check this in the sea object!!

# 4.6 Saving objects

```{r}
# Save plots as PNG
ggsave(
    filename = here("plots_and_figures", "heatmap_rho_cld_zma.png"),
    plot     = p_heatmap_rho,
    width    = 9, height = 3, dpi = 300
)

ggsave(
    filename = here("plots_and_figures", "zrho_by_dupmode_zma.png"),
    plot     = p_zrho,
    width    = 8, height = 3, dpi = 300
)

ggsave(
    filename = here("plots_and_figures", "expression_divergence_over_time_zma.png"),
    plot     = p_rho_ks_zma,
    width    = 9, height = 4, dpi = 300
)

ggsave(
    filename = here("plots_and_figures", "correlation_per_gene_family.png"),
    plot     = p_zrho_fam,
    width    = 9, height = 4, dpi = 300
)

ggsave(
    filename = here("plots_and_figures", "combined_expression_divergence.png"),
    plot     = p_zrho + p_zrho_fam + plot_annotation(tag_levels = "a"),
    width    = 11, height = 4, dpi = 300
)
```

