---
title: 'Chapter 3:'
author: "Woowon Jang"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(here)

set.seed(123) # for reproducibility
options(timeout = 1e6) # to download large data files

# Load helper functions
here::i_am("chapters/chapter2.Rmd")
source(here("products", "code", "utils.R"))

# Plot background
bg <- grid::linearGradient(colorRampPalette(c("gray90", "white"))(100))
```

# 2 Gene level analysis
## 2.1 Load packages
```{r}
library(rcompanion)
library(scuttle)
library(tidyverse)
library(SpatialExperiment)
library(SummarizedExperiment)
library(DESpace)
library(patchwork)
library(dplyr)
library(tibble)
library(Matrix)
library(stringr)
library(ggplot2)
```

Load data
- duplication data made in chapter2 ("duplication_info_filtered.rds")
- list of paralogs from uncorrected_sce.rds
```{r}
dup_list_valid <- readRDS(here("Duplication_data", "duplication_info_filtered.rds"))
sce_paralogs <- readRDS(here("chapters", "sce_paralogs.rds"))
```

Boxplots showing the distribution of Ka/Ks ratios across different gene duplication types
***Boxplot will not be used***
Excluded the Ka/Ks comparison as our core focus is on cell-type-specific expression dynamics.
```{}
output_dir <- here("products", "plots")
file_name <- "Ka_Ks_boxplot_valid.png"
full_path <- file.path(output_dir, file_name)

df_pairs <- dup_list_valid$pairs

my_colors <- c("DD" = "#9ecae1",
               "SD" = "#a1d99b",
               "dTRD" = "#fdbb84",
               "PD" = "#bcbddc",
               "rTRD" = "#d77fb3",
               "TD" = "#9e9ac8")

median_data <- df_pairs %>%
  group_by(type) %>%
  summarise(median_val = median(Ka_Ks, na.rm = TRUE)) %>%
  mutate(median_txt = sprintf("%.4f", median_val))

p <- ggplot(df_pairs, aes(x = type, y = Ka_Ks, fill = type)) +
  geom_boxplot(width = 0.6, outlier.alpha = 0.3) +
  
  scale_fill_manual(values = my_colors) +
  
  scale_y_continuous(breaks = seq(0, 6, by = 1)) +
  coord_cartesian(ylim = c(0, 6)) +

  theme_classic() +
  
  labs(title = "Ka/Ks Distribution by Type",
       x = "Duplication Type",
       y = "Ka/Ks Ratio") +
  
  theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      
      axis.title.x = element_text(size = 14, face = "bold", margin = margin(t = 10)),
      axis.title.y = element_text(size = 14, face = "bold", margin = margin(r = 10)),
      
      axis.text = element_text(color = "black", size = 12)
    )

ggsave(filename = full_path, plot = p, width = 8, height = 6, dpi = 300)
```

- counts: raw counts 
- logcounts: log-transformed normalized counts
```{r}
# Check assay names in the valid SpatialExperiment object
assayNames(sce_paralogs)
```
Confirmed that the reconstructed assay is in a sparse matrix format for memory efficiency
```{r}
# Check the reconstructed assay
assay(sce_paralogs, "logcounts")[1:5, 1:5]
```

# 2.2 Expression levels of duplicated genes
```{r}
# Define color palette for duplication modes
dup_pal <- c(
    SD = "#000000",  
    TD = "#E69F00",
    PD = "#56B4E9",
    rTRD = "#009E73",
    dTRD = "#F0E442",
    DD = "#0072B2"
)
```

Calculate the sum and mean expression level for all genes across samples.
```{r}
# Extract gene duplication information for valid genes
dup_genes <- dup_list_valid$genes
# dup_genes: 28902 x 2 (gene, type)

# sum across all samples
expr_sum <- rowSums(logcounts(sce_paralogs), na.rm = TRUE)

expr_sum_df <- tibble::tibble(
  gene = names(expr_sum),
  exp  = as.numeric(expr_sum)
)

sum_df <- expr_sum_df |>
  dplyr::inner_join(dup_genes, by = "gene") |>
  dplyr::mutate(
    species   = "zma",
    statistic = "sum",
    type      = factor(type, levels = names(dup_pal))
  )

# mean across all samples
expr_mean <- rowMeans(logcounts(sce_paralogs), na.rm = TRUE)

expr_mean_df <- tibble::tibble(
  gene = names(expr_mean),
  exp  = as.numeric(expr_mean)
)

mean_df <- expr_mean_df |>
  dplyr::inner_join(dup_genes, by = "gene") |>
  dplyr::mutate(
    species   = "zma",
    statistic = "mean",
    type      = factor(type, levels = names(dup_pal))
  )
```

Check the resulting data frames for sum expression levels
```{r}
nrow(sum_df)
head(sum_df)
```

Check the resulting data frames for mean expression levels
```{r}
nrow(mean_df)
head(mean_df)
```

Performs Kruskal-Wallis and Dunn's post-hoc tests (BH adjusted) to generate a Compact Letter Display (CLD) for comparing expression distributions across duplication types
```{r}
# Wrapper to compare distributions using Kruskal-Wallis + Dunn
#' @param df A data frame with values and variables.
#' @param var Character, name of column indicating variable levels.
#' @param value Numeric, values to be used in comparisons.
#'
#' @return A data frame with CLD for each level.

cld_kw_dunn <- function(df, var = "type", value = "csum") {
    
    # Perform Kruskal-Wallis test followed by a post-hoc Dunn's test
    dunn <- FSA::dunnTest(
        df[[value]] ~ df[[var]], method = "bh"
    ) |>
        purrr::pluck("res") |>
        arrange(Z)
    
    # Get compact letter display (CLD)
    cld <- rcompanion::cldList(
        comparison = dunn$Comparison,
        p.value = dunn$P.adj,
        threshold = 0.05
    )
    
    return(cld)
}
```

Check CLD for sum expression levels
```{r}
mean_comp <- cld_kw_dunn(mean_df, var = "type", value = "exp")
mean_comp
```

The violin plot compares mean expression by duplication mode with CLD labels for significance. For clarity, 522 outliers were excluded by capping the x-axis at 0.5.
```{r}
plot_mean <- ggplot(mean_df, aes(x = exp, y = type)) +   # <-- use mean_df, not mean_comp
  # Draw the violin plot
  geom_violin(aes(fill = type), show.legend = FALSE) +
  scale_fill_manual(values = dup_pal) +
  
  # Add the boxplot inside
  geom_boxplot(width = 0.1, outlier.color = "gray60", outlier.alpha = 0.5) +
  
  # Add CLD labels (rename Group -> type to match y-axis, set fixed x position)
  geom_label(
    data = mean_comp |>
      dplyr::rename(type = Group) |>
      dplyr::mutate(x = 0.4),
    aes(x = x, y = type, label = Letter),
    fill = "white", color = "black", label.size = 0.5, size = 5
  ) +
  
  # Set x-axis limit
  scale_x_continuous(limits = c(0, 0.5)) +
  
  # Set labels and titles
  labs(
    x = "Mean of log-transformed normalized counts", y = NULL,
    title = "Mean expression level and gene duplication mode",
    subtitle = "CLD from Kruskal-Wallis + post-hoc Dunn's test; P <0.05"
  ) +
  
  theme_classic() +
  theme(panel.background = element_rect(fill = bg))

# Display the plot
plot_mean

# Save the plot
ggsave(
  file.path(here("products", "plots"), "Zma_Mean_Expression_violin.png"),
  plot = plot_mean,
  width = 7,
  height = 5,
  dpi = 300
)
```

## 2.3 Expression breadths of duplicated genes

```{r}

```
