---
title: "chapter4"
output: html_document
execute: 
  echo: true
  eval: false
  warning: false
---

# Expression divergence between duplicate pairs

In this chapter, we will explore expression divergence between duplicate
pairs using different metrics to investigate if there is biased divergence
depending on the duplication mechanisms. 

First load required packages:
```{r, warning=FALSE, message=FALSE}
library(exdiva)
library(SingleCellExperiment)
library(SuperCell)
library(here)
```

! only run this block if you have the file on your computer !
```{r}
metacell_sce <- readRDS(here("chapters", "metacells.rds"))
```
Read in the SingleCellExperiment created in chapter1

We need to do metacell aggregation to reduce noise in the data.
We do this using the SuperCell package.

Create the metacells
```{r}
sce <- readRDS(here("chapters", "sce_paralogs.rds"))
sc <- SCimplify(
   X = assay(sce, "logcounts"),
   gamma = 30,
   cell.annotation = colData(sce)$Celltype_Organ
)
  
metacell_logcounts <- supercell_GE(
  ge = assay(sce, "logcounts"),
  groups = sc$membership,
  mode = "sum"
)
  
metacell_sce <- SingleCellExperiment(
  assays = list(logcounts = metacell_logcounts)
)
colData(metacell_sce)$celltype <- sc$sc.annotation$sc.cell.annotation
```

```{r}
saveRDS(metacell_sce, here("chapters", "metacell_sce.rds"))
saveRDS(sc, here("chapters", "sc.rds"))
```
We must add the colData to metacell_sce for further analysis.
Celltypes are aggregated per metacell by assigning each metacell the most common celltype among its   constituent cells. 

```{r}
sc$membership
```

