---
title: "Chapter ?: network-based analysis"
author: "Eli Galle"
date: "`r Sys.Date()`"
output: html_document
---

In this chapter, we will study the patterns of gene expression in paralogs using gene coexpression networks.


Load packages
```{r setup, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(tidyverse)
library(here)
library(BioNERO) # BiocManager::install("almeidasilvaf/BioNERO")
library(exdiva) # BiocManager::install("almeidasilvaf/exdiva")
library(HybridExpress)
```


For making figures work on hpc
```{r}
options(bitmapType = "cairo")
```

Setting aesthetics for figures
```{r}
# Plot background
bg <- grid::linearGradient(colorRampPalette(c("gray90", "white"))(100))

# Define color palette for duplication modes
dup_pal <- c(
    SD = "#000000",  
    TD = "#E69F00",
    PD = "#56B4E9",
    rTRD = "#009E73",
    dTRD = "#F0E442",
    DD = "#0072B2"
)
```

Loading in expression and paralog data and from chapter 2
```{r}
#sce <- readRDS(here("chapters", "sce_paralogs.rds"))
sce <- readRDS(here("chapters", "summarized_sce_paralogs.rds")) 
dup_info <- readRDS(here("Duplication_data", "duplication_info.rds"))
dup_genes <- dup_info$genes
dup_pairs <- dup_info$pairs
rm(dup_info)
```


## GCN inference

GCN inference is done using the package BioNERO. This package implements the WGCNA algorithm. 

Expression values are transformed into a pairwise similarities, calculated based on the pairwise gene correlations. An unsigned method treats negative correlations the same as positive by taking the absolute value. In a signed method the transformation is constructed a little different to only pick up on positive correlations.

Next, the similarity matrix is raised to a power beta to define the adjacency matrix. This is done to amplify distances. Values close to 1 will change relatively little while lower values will decrease. Using a higher beta results in a more scale-free topology (= most nodes have little edges, selected amount of a lot; desired characteristic in biological network). However, it also reduces mean connectivity.

The resulting network can then be analyzed using clustering methods to identify modules of genes that are adjacent.

```{r}
# to speed up calculations
# change nThreads depending on how many you have (parallel::detectCores())
WGCNA::allowWGCNAThreads(nThreads = 36)
```

GCN inference is done with log normalized counts
```{r}
mat <- sce
```


We will do some additional preprocessing steps
- filtering genes that are not (little) expressed
- adjustment for confounding artifacts
```{r}
mat <- remove_nonexp(mat, method = "median", min_exp=10)
```
mean of 1 also used

Optimal beta value is determined with the function SFT_fit()
```{r}
sft <- SFT_fit(mat, net_type = "signed hybrid", cor_method = "spearman")
```

Take a look at results
```{r}
sft$power
sft$plot
```



Infer GCN 
(Download from hpc)
```{r}
run_gcn = FALSE

if (run_gcn) {
    gcn <- exp2gcn(
    mat,
    net_type = "signed hybrid", # hybrid: sim = cor if cor>0 else 0
    SFTpower = 20,       
    cor_method = "spearman",   
    return_cormat = FALSE,      # whether you want the correlation matrix
    verbose = TRUE
  )
  saveRDS(gcn, "gcn.rds")

} else {
    gcn <- readRDS(here("chapters", "gcn.rds"))
}
```

Extract hubs; groups of highly co-expressed genes
```{r}
# Extract hubs,
final_df <- cbind(gcn$genes_and_modules, gcn$kIN) |>
  filter(Modules != "grey") # grey module is the trash bin module

# split in tables for each module; want every module in the gcn to be represented
# get top 10% 
hubs <- lapply(split(final_df, final_df$Modules), function(x) {
  hubs_vec <- x %>%
    arrange(-kWithin) %>%
    slice_head(n = round(nrow(x) * 0.1)) %>%
    pull(Genes)
  
  return(hubs_vec)
  }
)

# converting list with hub genes for each module into vector of all hub genes
# don't need to know from which module they were, so unname
hubs <- unname(unlist(hubs))
```

Summarize number of (paralog) hub genes per duplication mode, compare with background frequency.
```{r}
# count number of (paralog) hub genes per duplication mode
dup_genes[dup_genes$gene %in% hubs, ]$type %>% 
  table() %>%
  prop.table() %>% 
  round(3) * 100 #display as percentage

# count number of (paralog) genes per duplication mode (=background frequency)
dup_genes$type %>% 
  table() %>% 
  prop.table() %>% 
  round(3) * 100 #display as percentage
```
Statistical test of over representation

```{r}
ora_HubsDup <- HybridExpress::ora(genes = hubs,
    annotation = dup_genes,
    background = dup_genes$gene,
    max_setsize = 100000,
    alpha = 1 # want all p-values for plot
  )
ora_HubsDup
```

Visualize with plot
```{r}
# reorder the df to the order in the color pallete for the duplication modes
ora_HubsDup <- ora_HubsDup[match(names(dup_pal), ora_HubsDup$term),]

figure_ora <- ora_HubsDup %>%
  mutate(
    term = factor(term, levels = c("SD",
                                   "TD",
                                   "PD",
                                   "rTRD",
                                   "dTRD",
                                   "DD")),
    logP = -log10(padj),
    significant = ifelse(padj < 0.05, TRUE, FALSE),
    symbol = case_when(
      padj <= 0.05 & padj > 0.01 ~ "*",
      padj <= 0.01 & padj > 0.001 ~ "**",
      padj <= 0.001 ~ "***",
      TRUE ~ ""
    )
  ) %>% 
  ggplot(aes(x=genes, y=term)) +
    geom_point(
      aes(fill=term, size = logP, alpha = significant),
      color = "gray20", pch = 21
    ) +
    scale_size(range = c(4, 9)) +
    scale_alpha_manual(values = c(0.3, 1)) +
    scale_fill_manual(values = dup_pal) +
    geom_text(aes(label = symbol), vjust = -0.5, size = 5) +
    scale_x_continuous(limits = c(0, 900)) +
    theme_classic() +
    guides(fill = "none", alpha = "none") +
    labs(
        title = "Duplication mode of hub genes in coexpression networks",
        subtitle = "Bright dots with asterisks indicate significant overrepresentation; P <0.05",
        x = "Number of hub genes", y = NULL,
        size = expression(-Log[10] ~ italic(P))
    )

figure_ora
```
```{r}
ggsave(here("plots_and_figures", "figure_ORA_DupMode_HubGenes.png"), 
       figure_ora,
       width = 6,
       height = 5)
```


TODO

scater:plotReducedDim to plot 2 paralogs on the general 

QUESTIONS

Resolved:
- GCN inference should probably be with expression of every gene (not just the paralogs) --> not necessary
- metalcells --> seurat package hdWGCNA for metacell or metacells on python (convert rds files with zellkonverter)
- preprocessing see vignete BioNERO, particularly the PCA_correct (see paper in ?PC_correction) + no prepro in Fabricio paper?
--> filter for low but no PCA_correct

New:
- sft results
- why would you have max_setsize in HybridExpress::ora


  





