---
title: "Chapter ?: network-based analysis"
author: "Eli Galle"
date: "`r Sys.Date()`"
output: html_document
---

In this chapter, we will study the patterns of gene expression in paralogs using gene coexpression networks.


Load packages
```{r setup, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(tidyverse)
library(here)
library(BioNERO)
library(exdiva) # had to install this one from GitHub, not available on Bioconductor
```


For reproducibility
```{r}
set.seed(604)
```

Setting aesthetics for figures
```{r}
# Plot background
bg <- grid::linearGradient(colorRampPalette(c("gray90", "white"))(100))

# Define color palette for duplication modes
dup_pal <- c(
    SD = "#000000",  
    TD = "#E69F00",
    PD = "#56B4E9",
    rTRD = "#009E73",
    dTRD = "#F0E442",
    DD = "#0072B2"
)
```

Loading in expression and paralog data and from chapter 2
```{r}
#sce <- readRDS(here("chapters", "sce_paralogs.rds"))
sce <- readRDS(here("chapters", "summarized_sce_paralogs.rds")) # placeholder for metacells
dup_info <- readRDS(here("Duplication_data", "duplication_info.rds"))
dup_genes <- dup_info$genes
dup_pairs <- dup_info$pairs
rm(dup_info)
```


## Metacell aggregation

Due to the high sparsity of scRNA data, we will aggregate cells into metacells before GCN inference.



## GCN inference

GCN inference is done using the package BioNERO. This package implements the WGCNA algorithm. 

Expression values are transformed into a pairwise similarities, calculated the pairwise gene correlations. An unsigned method treats negative correlations the same as positive by taking the absolute value. In a signed method the transformation is constructed a little different to only pick up on positive correlations.

Next, the similarity matrix is raised to a power beta to define the adjacency matrix. This is done to amplify distances. Values close to 1 will change relatively little while lower values will decrease. Using a higher beta results in a more scale-free topology (= most nodes have little edges, selected amount of a lot; desired characteristic in biological network). However, it also reduces mean connectivity.

The resulting network can then be analyzed using clustering methods to identify modules of genes that are adjacent.

```{r}
# to speed up calculations
# change nThreads depending on how many you have (parallel::detectCores())
WGCNA::allowWGCNAThreads(nThreads = 10)
```

GCN inference is done with log normalized counts
```{r}
mat <- as.matrix(assay(sce, "logcounts.sum"))

mat <- mat[1:1000, ] # temporary for testing
```


Optimal beta value is determined with the function SFT_fit()
```{r}
sft <- SFT_fit(mat, net_type = "signed hybrid", cor_method = "pearson")
```

Take a look at results
```{r}
sft$power
sft$plot
```
Detected optimal beta = 18

Infer GCN
```{r}
gcn <- exp2gcn(
  mat,
  net_type = "signed hybrid",
  SFTpower = sft$power,       # use optimal beta from SFT_fit function
  cor_method = "spearman",   
  return_cormat = FALSE,      # whether you want the correlation matrix
  verbose = TRUE
)
```


Extract hubs; groups of highly co-expressed genes
```{r}
hubs <- get_hubs_gcn(mat, gcn)
```



QUESTIONS

- GCN inference should probably be with expression of every gene
- simpleWarning in eval(xpr, envir = envir): Some correlations are NA in block 28843 : 29469 . * 195 when running SFT_fit






