---
title: "chapter2"
author: "Woowon Jang"
date: "`r Sys.Date()`"
output: html_document
---

# 1 Environment settings
## 1.1 Load packages
```{r}
library(here)
library(tidyverse)
library(SpatialExperiment)
library(SummarizedExperiment)
library(DESpace)
library(patchwork)
library(dplyr)
library(tibble)
library(Matrix)
library(stringr)
library(ggplot2)
library(SingleCellExperiment)

set.seed(123) # for reproducibility
options(timeout = 1e6) # to download large data files

# Load helper functions
here::i_am("chapters/chapter2.Rmd")
source(here("products", "code", "utils.R"))

# Plot background
bg <- grid::linearGradient(colorRampPalette(c("gray90", "white"))(100))
```

## 1.2 Load duplication data
```{r}
# Load duplicate pairs and genes
dup_list <- readRDS(
    here("products", "results", "dup_list.rds")
)

# Define color palette for duplication modes
dup_pal <- c(
    SD = "#000000",  
    TD = "#E69F00",
    PD = "#56B4E9",
    rTRD = "#009E73",
    dTRD = "#F0E442",
    DD = "#0072B2"
)
```

Overview of the duo_list data
```{r}
head(dup_list[["zma"]]$pairs, 5)
head(dup_list[["zma"]]$genes, 5)
```

```{r}
names(dup_list[["zma"]])

names(dup_list[["zma"]]$pairs)
names(dup_list[["zma"]]$genes)
```

## 1.3 Load Zeo Maize data
```{r}
# Load `SpatialExperiment` objects
zma_spe <- readRDS(here("products", "results", "corrected_RedDim.rds"))
```

## 1.4 Check gene ID compatibility
```{r}
# V4 zma_spe
# Extract the numbers at the end
user_genes <- rownames(zma_spe)
# Extract the ID that starts with Zm00
user_genes <- user_genes[grep("Zm00", user_genes)]
# Remove Zm00001d pattern and get the second part
user_nums <- gsub("Zm00001d", "", user_genes)

# V5 dup_list
# Extract the numbers at the end
list_genes <- dup_list[["zma"]]$genes$gene
# Remove Zm00001eb pattern and get the second part
list_nums <- gsub("Zm00001eb", "", list_genes)

# Check the intersects between dup_list and zma_spe
common_nums <- intersect(user_nums, list_nums)

cat("zma_spe(v4) number:", length(user_nums), "\n")
cat("dup_list(v5) number:", length(list_nums), "\n")
cat("overlap:", length(common_nums), "\n")
```

Extract Gene ID from zme_spe
```{}
output_dir <- here("products", "tables")
txt_filename <- "zma_gene_ids.txt"
txt_full_path <- file.path(output_dir, txt_filename)

# Extract Gene ID
gene_ids <- rownames(zma_spe)

# Save as .txt file
write.table(gene_ids, file = txt_full_path, 
            quote = FALSE, 
            row.names = FALSE, 
            col.names = FALSE)
```

Open and adjust the converted file
e.g. Zm00001d001763.1 -> Zm00001d001763
```{r}
# Clean IDs in the mapping table (Remove version suffixes like '.1')
zma_id_raw <- read.csv(here("products", "tables", "zma_ID_converter.txt"), 
                     check.names = FALSE, 
                     stringsAsFactors = FALSE)

id_cleaned <- zma_id_raw %>%
  select(`Old stable ID`, `New stable ID`) %>%
  filter(grepl("^\\s*Zm", `Old stable ID`)) %>%
  mutate(
    Old_Cleaned = sub("\\..*", "", `Old stable ID`),
    New_Cleaned = sub("\\..*", "", `New stable ID`)
  )

head(id_cleaned[, c("Old stable ID", "Old_Cleaned")])
```

```{r}
# Extract IDs from your data objects (ensure these are loaded)
spe_ids <- rownames(zma_spe)
dup_ids <- dup_list[["zma"]]$genes$gene

spe_ids_clean <- sub("\\..*", "", spe_ids)

# Check existence using the CLEANED IDs
final_match <- id_cleaned %>%
  mutate(
    # Delete empty space 
    Old_Cleaned = trimws(Old_Cleaned),
    New_Cleaned = trimws(New_Cleaned),
    
    match_spe = Old_Cleaned %in% spe_ids_clean,
    match_dup = New_Cleaned %in% dup_ids,
    valid = match_spe & match_dup
  )

# Count valid matches
cat("\nNumber of valid (mappable) genes:", sum(final_match$valid), "\n")

# Save final_match object as RDS
output_path <- here("products", "results", "final_match.rds")
saveRDS(final_match, file = output_path)
```

```{r}
# Load the saved final_match data
final_match <- readRDS(here("products", "results", "final_match.rds"))

# Check the loaded data
head(final_match)

valid_matches <- final_match %>% 
  filter(valid == TRUE)

valid_spe_ids <- valid_matches$Old_Cleaned

zma_spe_valid <- zma_spe[valid_spe_ids, ]
rownames(zma_spe_valid) <- valid_matches$New_Cleaned

nrow(valid_matches)
```

```{r}
# Create zma_spe_valid
valid_spe_ids <- valid_matches$Old_Cleaned
zma_spe_valid <- zma_spe[valid_spe_ids, ]
rownames(zma_spe_valid) <- valid_matches$New_Cleaned

# Create dup_list_valid
dup_list_valid <- dup_list
valid_dup_ids <- valid_matches$New_Cleaned

# Filter the genes data frame
dup_list_valid[["zma"]]$genes <- dup_list[["zma"]]$genes %>%
  filter(trimws(gene) %in% valid_dup_ids)

# Filter the pairs data frame
dup_list_valid[["zma"]]$pairs <- dup_list[["zma"]]$pairs %>%
  filter(trimws(dup1) %in% valid_dup_ids & trimws(dup2) %in% valid_dup_ids)
```

```{r}
head(rownames(zma_spe_valid))
```

```{}
# Save data
saveRDS(zma_spe_valid, file = here("products", "results", "zma_spe_valid.rds"))
saveRDS(dup_list_valid, file = here("products", "results", "dup_list_valid.rds"))
```

```{r}
zma_spe_valid <- readRDS(here("products", "results", "zma_spe_valid.rds"))
dup_list_valid <- readRDS(here("products", "results", "dup_list_valid.rds"))
```

```{r}
# Verification
cat("Original zma_spe genes:", nrow(zma_spe), "\n")
cat("Valid zma_spe genes:   ", nrow(zma_spe_valid), "\n\n")

cat("Original dup_list pairs:", nrow(dup_list[["zma"]]$pairs), "\n")
cat("Valid dup_list pairs:   ", nrow(dup_list_valid[["zma"]]$pairs), "\n")
```

Boxplots showing the distribution of Ka/Ks ratios across different gene duplication types
```{r}
output_dir <- here("products", "plots")
file_name <- "Ka_Ks_boxplot_valid.png"
full_path <- file.path(output_dir, file_name)

df_pairs <- dup_list_valid[["zma"]]$pairs

my_colors <- c("DD" = "#9ecae1",
               "SD" = "#a1d99b",
               "dTRD" = "#fdbb84",
               "PD" = "#bcbddc",
               "rTRD" = "#d77fb3",
               "TD" = "#9e9ac8")

median_data <- df_pairs %>%
  group_by(type) %>%
  summarise(median_val = median(Ka_Ks, na.rm = TRUE)) %>%
  mutate(median_txt = sprintf("%.4f", median_val))

p <- ggplot(df_pairs, aes(x = type, y = Ka_Ks, fill = type)) +
  geom_boxplot(width = 0.6, outlier.alpha = 0.3) +
  
  scale_fill_manual(values = my_colors) +
  
  scale_y_continuous(breaks = seq(0, 6, by = 1)) +
  coord_cartesian(ylim = c(0, 6)) +

  theme_classic() +
  
  labs(title = "Ka/Ks Distribution by Type",
       x = "Duplication Type",
       y = "Ka/Ks Ratio") +
  
  theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      
      axis.title.x = element_text(size = 14, face = "bold", margin = margin(t = 10)),
      axis.title.y = element_text(size = 14, face = "bold", margin = margin(r = 10)),
      
      axis.text = element_text(color = "black", size = 12)
    )

ggsave(filename = full_path, plot = p, width = 8, height = 6, dpi = 300)
```

# 2 Gene level analysis
## 2.1 Load packages
```{r}
library(rcompanion)
library(scuttle)
```


```{r}
# Check assay names in the valid SpatialExperiment object
assayNames(zma_spe_valid)
```

```{r}
# Check the reconstructed assay
assay(zma_spe_valid, "logcounts")[1:5, 1:5]
```

# 2.2 Expression levels of duplicated genes
```{r}
dup_genes <- dup_list_valid[["zma"]]$genes
# dup_genes: 28902 x 2 (gene, type)

# sum
expr_sum <- rowSums(logcounts(zma_spe_valid), na.rm = TRUE)

expr_sum_df <- tibble::tibble(
  gene = names(expr_sum),
  exp  = as.numeric(expr_sum)
)

sum_df <- expr_sum_df |>
  dplyr::inner_join(dup_genes, by = "gene") |>
  dplyr::mutate(
    species   = "zma",
    statistic = "sum",
    type      = factor(type, levels = names(dup_pal))
  )

nrow(sum_df)
head(sum_df)

# mean
expr_mean <- rowMeans(logcounts(zma_spe_valid), na.rm = TRUE)

expr_mean_df <- tibble::tibble(
  gene = names(expr_mean),
  exp  = as.numeric(expr_mean)
)

mean_df <- expr_mean_df |>
  dplyr::inner_join(dup_genes, by = "gene") |>
  dplyr::mutate(
    species   = "zma",
    statistic = "mean",
    type      = factor(type, levels = names(dup_pal))
  )

nrow(mean_df)
head(mean_df)
```
```{r}
# Combine all `SpatialExperiment` objects in a single list
sum_combined <- sum_df |>
  dplyr::mutate(
    csum    = exp,
    species = "Zma"
  ) |>
  dplyr::select(gene, type, species, csum) |>
  dplyr::distinct(gene, .keep_all = TRUE)

# Get mean of gene expression levels across all samples
mean_combined <- mean_df |>
  dplyr::mutate(
    cmean   = exp,
    species = "Zma"
  ) |>
  dplyr::select(gene, type, species, cmean) |>
  dplyr::distinct(gene, .keep_all = TRUE)
```

```{r}
# Compare distros and get CLDs
## Sum
sum_clds <- cld_kw_dunn(
  sum_combined,
  var   = "type",
  value = "csum"
) |>
  dplyr::mutate(
    species = factor("Zma", levels = c("Zma")),
    x       = max(sum_combined$csum, na.rm = TRUE) * 0.9
  ) |>
  dplyr::rename(type = Group)

#### Mean
mean_clds <- cld_kw_dunn(
  mean_combined,
  var   = "type",
  value = "cmean"
) |>
  dplyr::mutate(
    species = factor("Zma", levels = c("Zma")),
    x       = max(mean_combined$cmean, na.rm = TRUE) * 0.9
  ) |>
  dplyr::rename(type = Group)
```

```{r}
p_sum_combined <- ggplot(sum_combined, aes(x = csum, y = type)) +
  geom_violin(aes(fill = type), show.legend = FALSE) +
  scale_fill_manual(values = dup_pal) +
  geom_boxplot(width = 0.1, outlier.color = "gray60", outlier.alpha = 0.5) +
  geom_label(
    data = sum_clds,
    aes(x = x, y = type, label = Letter)
  ) +
  facet_wrap(~ species, nrow = 1, scales = "free_x") +
  ggh4x::facetted_pos_scales(x = list(
    scale_x_continuous(
      limits = c(0, max(sum_combined$csum, na.rm = TRUE) * 1.05),
      labels = scales::unit_format(unit = "K", scale = 1e-3)
    )
  )) +
  labs(
    x = "Sum of log-transformed normalized counts", y = NULL,
    title = "Total expression levels and gene duplication mode (Zma)",
    subtitle = "CLD from Kruskal-Wallis + post-hoc Dunn's test; P < 0.05"
  ) +
  theme_classic() +
  theme(panel.background = element_rect(fill = bg))

p_mean_combined <- ggplot(mean_combined, aes(x = cmean, y = type)) +
  geom_violin(aes(fill = type), show.legend = FALSE) +
  scale_fill_manual(values = dup_pal) +
  geom_boxplot(width = 0.1, outlier.color = "gray60", outlier.alpha = 0.5) +
  geom_label(
    data = mean_clds, 
    aes(x = x, y = type, label = Letter)
  ) +
  facet_wrap(~ species, nrow = 1, scales = "free_x") +
  ggh4x::facetted_pos_scales(x = list(
    scale_x_continuous(
      limits = c(0, max(mean_combined$cmean, na.rm = TRUE) * 1.05)
    )
  )) +
  labs(
    x = "Mean of log-transformed normalized counts", y = NULL,
    title = "Mean expression level and gene duplication mode (Zma)",
    subtitle = "CLD from Kruskal-Wallis + post-hoc Dunn's test; P < 0.05"
  ) +
  theme_classic() +
  theme(panel.background = element_rect(fill = bg))

```

```{r}
output_dir <- here("products", "plots")

ggsave(
  file.path(output_dir, "Zma_sum_combined.png"),
  plot   = p_sum_combined,
  width  = 10,
  height = 6,
  dpi    = 300
)

ggsave(
  file.path(output_dir, "Zma_mean_combined.png"),
  plot   = p_mean_combined,
  width  = 10,
  height = 6,
  dpi    = 300
)
```
# 3.2 Expression breadths of duplicated genes
```{r}
#' Calculate the proportion of non-zero spots for each gene by cell type
#'
#' @param spe A SingleCellExperiment or SpatialExperiment object.
#' @param cell_type Character, name of the column with cell type information.
#'
#' @return A data frame with variables `gene`, `cell_type`, and `prop_detected`.
get_prop_detected <- function(spe,
                              cell_type  = "Celltype",
                                assay_name = "logcounts"
                                ) {
  
  mat <- SummarizedExperiment::assay(spe, assay_name)
  
  ct <- as.character(SummarizedExperiment::colData(spe)[[cell_type]])
  if (is.null(ct)) {
    stop("Column '", cell_type, "' not found in colData(spe).")
  }
  
  res_list <- lapply(unique(ct), function(cti) {
    idx  <- which(ct == cti)
    prop <- rowMeans(mat[, idx, drop = FALSE] != 0)
    
    tibble::tibble(
      gene          = rownames(mat),
      cell_type     = cti,
      prop_detected = as.numeric(prop)
    )
  })
  
  dplyr::bind_rows(res_list)
}

prop_detected_zma <- get_prop_detected(
  spe       = zma_spe_valid,
  cell_type = "Celltype"   # colData(zma_spe_valid)$Celltype
)

head(prop_detected_zma)

output_dir <- here("products", "results")
saveRDS(prop_detected_zma, file = file.path(output_dir, "prop_detected_zma.rds"))
```

```{r}
# Calculate absolute expression breadth
eb_zma <- prop_detected_zma |>
  dplyr::group_by(gene, cell_type) |>
  dplyr::mutate(mean_prop = mean(prop_detected, na.rm = TRUE)) |>
  dplyr::ungroup() |>
  dplyr::filter(mean_prop >= 0.01) |>
  dplyr::distinct(gene, cell_type, .keep_all = TRUE) |>
  dplyr::count(gene, name = "breadth") |> 
  dplyr::inner_join(dup_list_valid[["zma"]]$genes, by = "gene") |>
  dplyr::mutate(
    species = "Zma",
    type    = factor(type, levels = names(dup_pal))
  )

head(eb_zma)
```

```{r}
# Test for differences in expression breadth by duplication mode
eb_test <- cld_kw_dunn(
  eb_zma,
  var   = "type",
  value = "breadth"
) |>
  dplyr::mutate(
    species = factor("Zma", levels = "Zma"),
    x = max(eb_zma$breadth, na.rm = TRUE) * 0.9
  ) |>
  dplyr::rename(type = Group)
```

```{r}
# Plot distributions of absolute expression breadths
p_eb_zma <- ggplot(eb_zma, aes(x = breadth, y = type)) +
  geom_violin(aes(fill = type), show.legend = FALSE) +
  scale_fill_manual(values = dup_pal) +
  geom_boxplot(width = 0.1, outlier.color = "gray60", outlier.alpha = 0.5) +
  geom_label(
    data = eb_test,
    aes(x = x, y = type, label = Letter)
  ) +
  labs(
    x = "Expression breadth (number of cell types)",
    y = NULL,
    title = "Expression breadth and gene duplication mode (Zma)",
    subtitle = "CLD from Kruskal-Wallis + post-hoc Dunn's test; P < 0.05"
  ) +
  theme_classic() +
  theme(panel.background = element_rect(fill = bg))

ggsave(
  file.path(here("products", "plots"), "Zma_expression_breadth.png"),
  plot   = p_eb_zma,
  width  = 8,
  height = 5,
  dpi    = 300
)
```

# 3.3 Spatial variability of duplicated genes
```{r}
assayNames(zma_spe_valid)

assay(zma_spe_valid, "counts") <- assay(zma_spe_valid, "logcounts")

assayNames(zma_spe_valid)

```

```{r}
# Define helper function to identify SVGs with DESpace
get_svg <- function(spe, spatial_cluster = "Celltype") {
  
  res <- DESpace_test(
    spe              = spe,
    spatial_cluster  = spatial_cluster,
    replicates       = FALSE,
    min_counts       = 1,
    min_non_zero_spots = 5
  )
  gc()
  
  res_df <- res$gene_results |>
    as.data.frame() |>
    tibble::rownames_to_column("gene_id") |>
    dplyr::filter(!is.na(FDR), FDR <= 0.05)
  
  return(res_df)
}

colData(zma_spe_valid)$Celltype   <- factor(colData(zma_spe_valid)$Celltype)
colData(zma_spe_valid)$Orig.ident <- factor(colData(zma_spe_valid)$Orig.ident)
colData(zma_spe_valid)$Condition  <- factor(colData(zma_spe_valid)$Condition)

res <- dsp_test(
  spe = zma_spe_valid,
  cluster_col = "Celltype",
  sample_col = "Orig.ident",
  condition_col = "Condition",
  test = "QLF",
  min_counts = 1,
  min_non_zero_spots = 5,
  verbose = TRUE
)
```

```{r}
# Get svgs
get_svg <- function(spe, spatial_cluster = "Celltype", fdr_cutoff = 0.05) {

  res <- DESpace::svg_test(
    spe = spe,
    cluster_col = spatial_cluster,
    min_counts = 1,
    min_non_zero_spots = 5
  )

  res_df <- res$gene_results |>
    as.data.frame() |>
    tibble::rownames_to_column("gene_id") |>
    dplyr::filter(!is.na(FDR), FDR <= fdr_cutoff)

  res_df
}

# Identify SVGs
svgs <- list(
  zma = get_svg(zma_spe_valid, spatial_cluster = "Celltype")
)


```

```{r}
# Define helper function to perform ORA for duplication modes
ora_dupmode <- function(svg_df, dup_df) {
  df <- HybridExpress::ora(
    genes       = svg_df$gene_id,
    annotation  = as.data.frame(dup_df),
    background  = dup_df$gene,
    min_setsize = 2,
    max_setsize = 1e8
  )
  return(df)
}

# Zma duplication annotation
dup_zma_genes <- dup_list_valid[["zma"]]$genes  # columns: gene, type

ora_svg_dup_zma <- lapply(
  svgs[["zma"]],
  ora_dupmode,
  dup_df = dup_zma_genes
) |>
  dplyr::bind_rows(.id = "sample") |>
  dplyr::mutate(species = "Zma") |>
  dplyr::select(
    species,
    sample,
    type = term,
    genes,
    all,
    padj
  )

head(ora_svg_dup_zma)
```
