---
title: "chapter4"
output: html_document
execute: 
  echo: true
  eval: false
  warning: false
---

# Expression divergence between duplicate pairs

In this chapter, we will explore expression divergence between duplicate
pairs using different metrics to investigate if there is biased divergence
depending on the duplication mechanisms. 

First load required packages:
```{r}
library(exdiva)
library(SingleCellExperiment)
library(SuperCell)
```


Read in the SingleCellExperiment created in chapter1:
```{r}
sce <- readRDS("uncorrected_sce_RedDim.rds")
```


We need to do metacell aggregation to reduce noise in the data.
We do this using the SuperCell package

Create the metacells
```{r}
sc <- SCimplify(
  X = assay(sce, 'logcounts'),
  gamma = 30,  # proportion of number of single cells in the initial dataset to the number of metacells in the final dataset
)
```
sc$membership gives the metacell ID for each cell

```{r}
metacell_counts <- supercell_GE(
  ge = assay(sce, "counts"),
  groups = sc$membership,
  mode = "sum"
)

metacell_sce <- SingleCellExperiment(
  assays = list(counts = metacell_counts)
)
```

