---
title: "Chapter_5"
author: "Emmeline Van Rijswijck"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Frequency of genes in each divergence class originating from each duplication mode

We will start from the data obtained in chapter 2; duplication_info_filtered (and summarized_sce_paralogs)

# Reading in the data

```{r}
duplication_info_filtered <- readRDS("../Duplication_data/duplication_info_filtered.rds")
duplication_info_filtered$genes
```

# Calculate the relative expression breadth

The relative expression breadth REBA for gene A is defined as the proportion of cell types in which gene A is expressed relative to the total number of cell types in which either gene A or gene B is expressed.

To calculate relative expression breadth, a dataset is required that indicates which genes are expressed in which cell types, typically represented as a matrix where each gene is scored for expression across all analyzed cell types; the breadth is then computed as the fraction of cell types in which a gene is expressed.

Let's construct a matrix containing the following variables:
- Gene 
- Duplication mode
- Location of expression 

Metacells are used to reduce technical noise and dropout effects inherent to single-cell RNA sequencing data by aggregating transcriptionally similar cells, resulting in more robust expression profiles that allow a more reliable estimation of gene expression breadth across cell types.

```{r}
metacells <- readRDS('../Chapters/metacells.rds')
names(assay(metacells))
```
#tissue_organ gebruiken!!! niet celltype!!!!!!! ale dus om REB te berekenen 




