---
title: "chapter4"
output: html_document
execute: 
  echo: true
  eval: false
  warning: false
---

# Expression divergence between duplicate pairs

In this chapter, we will explore expression divergence between duplicate
pairs using different metrics to investigate if there is biased divergence
depending on the duplication mechanisms. 

First load required packages:
```{r, warning=FALSE, message=FALSE}
library(exdiva)
library(SingleCellExperiment)
library(SuperCell)
library(here)
library(dplyr)
```

Read in the SingleCellExperiment created in chapter1

We need to do metacell aggregation to reduce noise in the data.
We do this using the SuperCell package.

# 1 Create the metacells

! You only have to run this once, if you already created the metacell_sce.rds file: go to 4.2

```{r}
sce <- readRDS(here("chapters", "sce_paralogs.rds"))
```

```{r}
sc <- SCimplify(
   X = assay(sce, "logcounts"),
   gamma = 30,
   cell.annotation = colData(sce)$Celltype_Organ
)
```

```{r}
length(sc$membership)
max(sc$membership)
```
We have aggregated 228 359 singlecells into 18 196 metacells

```{r}
metacell_logcounts <- supercell_GE(
  ge = assay(sce, "logcounts"),
  groups = sc$membership,
  mode = "sum"
)
  
metacell_sce <- SingleCellExperiment(
  assays = list(logcounts = metacell_logcounts)
)
colData(metacell_sce)$celltype <- sc$SC.cell.annotation. ## Add annotation celltype_organ to the metacells
```

```{r}
saveRDS(metacell_sce, here("chapters", "metacell_sce.rds"))
```

# 2. Comparing distributions of gene-gene correlations by duplication mode

read in the metacells
```{r}
metacell_sce <- readRDS(here("chapters", "metacell_sce.rds"))
```

```{r}
paralog_pairs <- readRDS(here("Duplication_data", "duplication_info_filtered.rds"))

paralog_pairs$pairs %>%
     distinct(peak_name)
```
```{r}
paralog_pairs$pairs <- paralog_pairs$pairs[!is.na(paralog_pairs$pairs$peak_name), ]
```


```{r}
# Define color palette for duplication modes
dup_pal <- c(
    SD = "#000000",  
    TD = "#E69F00",
    PD = "#56B4E9",
    rTRD = "#009E73",
    dTRD = "#F0E442",
    DD = "#0072B2"
)
```

```{r}
# Create a dataframe containing only the duplicated pairs
dups <- as.data.frame(paralog_pairs$pairs[, c("dup1", "dup2", "type")])

# calculate the Spearman's correlation coëfficiënt using the exdiva package
se_simple <- SummarizedExperiment(
  assays = list(logcounts = as.matrix(logcounts(metacell_sce)))
)

sp_coeff <- calculate_cor(
  se = se_simple,
  dups = dups,
  assay_name = "logcounts",
  method = "spearman"
)
```

Remove pairs where correlation is NA
```{r}
sp_coeff <- sp_coeff %>%
     filter(!is.na(cor))
```


Next, for each sample, we will test for significant differences by 
duplication mode using Kruskal-Wallis tests followed by a post-hoc Dunn's test. 
To account for differences in ages between gene pairs, we will do this 
separately for each age group.

```{r}
# Ad all info to the correlation dataframe
sp_coeff_age <- dplyr::inner_join(paralog_pairs$pairs, sp_coeff, by = c("dup1", "dup2"))

```



```{r}
# Credits to Fabricio Almeida Silva 
# Compare distros
#' Wrapper function to compare distros with Kruskal-Wallis + Dunn
cld_kw_dunn <- function(df, var = "type.x", value = "csum") {
    
    # Perform Kruskal-Wallis test followed by a post-hoc Dunn's test
    dunn <- FSA::dunnTest(
        df[[value]] ~ df[[var]], method = "bh"
    ) |>
        purrr::pluck("res") |>
        arrange(Z)
    
    # Get compact letter display (CLD)
    cld <- rcompanion::cldList(
        comparison = dunn$Comparison,
        p.value = dunn$P.adj,
        threshold = 0.05
    )
    
    return(cld)
}

compare_cors <- function(rho_df) {
    
    # Add a variable indicating sample + Ks peak
    rho_df$id <- paste0(rho_df$sample, "_", rho_df$peak)
    ids <- unique(rho_df$id)
    
    # Perform tests and get CLD for each sample-peak
    cld_df <- Reduce(rbind, lapply(ids, function(x) {
        
        fdf <- rho_df[rho_df$id == x, ]
        
        # Perform Kruskal-Wallis test followed by a post-hoc Dunn's test
        cld <- cld_kw_dunn(fdf, value = "cor")
        cld$sample <- unique(fdf$sample)
        cld$peak <- unique(fdf$peak)
        cld$peak_name <- unique(fdf$peak_name)
        
        return(cld)
    }))
    
    
    
    return(cld_df)
}
```

```{r, warning= FALSE, message=FALSE}
comp_cors <- compare_cors(sp_coeff_age) |> mutate(species = "Zma")
```
!!! Check what type should be??


```{r}
# Plot heatmap with CLDs
p_heatmap_rho <- comp_cors |>
    mutate(Group = factor(Group, levels = rev(names(dup_pal)))) |>
    ggplot(aes(x = sample, y = Group, fill = Letter)) +
    geom_tile(color = "gray90", linewidth = 0.8, alpha = 0.7) +
    geom_text(aes(label = Letter), size = 3) +
    ggh4x::facet_nested(~ species + peak_name, scales = "free", space = "free") +
    theme_classic() +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    ) +
    labs(
        title = "Differences in gene-gene correlations by duplication mode",
        subtitle = "CLD from Kruskal-Wallis + post-hoc Dunn's test; P <0.05",
        x = "Samples", y = "Mode"
    ) +
    theme(
        legend.position = "none"
    ) +
    ggsci::scale_fill_npg()

p_heatmap_rho
```

