---
title: "Chapter 2: Acquiring Duplication Information"
author: "Woowon Jang, Eli Galle"
date: "`r Sys.Date()`"
output: html_document
---

In this chapter we will annotate the genes from our single cell experiment objects with duplication information (duplication mode, age (group) and paralog pairs)

# 1 Environment settings
## 1.1 Load packages

```{r, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(SummarizedExperiment)
library(DESpace)
library(patchwork)
library(Matrix)
library(SingleCellExperiment)
```


## 1.2 Load duplication data

zma_allgenes.tsv contains a list of all paralog genes of zea mays and their duplication type.
zma_allpairs.tsv contains a list of all the paralog **pairs**, as well as info on the duplication type, age (ks/ka) and if possible, the age group (there were two peaks of gene duplication in the evolutionary history of zea mays corresponding to two WGD events)

```{r, message=FALSE}
paralog_genes <- read_tsv(here("Duplication_data", "zma_allgenes.tsv"))
paralog_pairs <- read_tsv(here("Duplication_data", "zma_allpairs.tsv"))
```
Combine these 2 into 1 object
```{r}
duplication_info <- list(genes = paralog_genes,
                         pairs = paralog_pairs)
rm(paralog_genes, paralog_pairs)
duplication_info
```

Overview of the duplication_info data
```{r}
names(duplication_info)
names(duplication_info$pairs)
names(duplication_info$genes)
```
```{r}
head(duplication_info$pairs, 5)
head(duplication_info$genes, 5)
```

## 1.3 Load Zea Mays data

Load in the single cell sequencing data obtained from chapter 1. 
```{r}
summarized_sce <- readRDS(here("chapters","summarized_sce.rds"))
```

## 1.4 Check gene ID compatibility

Extract gene IDs used in the single cell experiment
```{r}
sceIDs <- rownames(summarized_sce)
head(sceIDs)
```
It contains a mixed annotation set with different ID types:
- Zm00001dxxxxx: these are standard nuclear protein-coding genes, the d
- ENSRNAxxxxxxxxx: RNA gene IDs, often non coding RNAs, but also predicted transcripts
- ZeamMpxxx: mitochondrial protein genes
- ZeamMtxxx: mitochondrial tRNA genes
- ZemaCpxxx: chloroplast genes
- ZemaCtxxx: chloroplast tRNA or other plastid non-coding genes

The d in Zm00001dxxxxx means that the IDs are in genome version V4

We will only analyse nuclear protein-coding genes, so IDs starting with Zm00 are extracted

```{r}
sceIDs <- sceIDs[grep("Zm00", sceIDs)]
length(sceIDs)
```
Retain 41856 genes (from the original 67300)
```{r}
# Remove Zm00001d pattern and get the second part
sce_nums <- gsub("Zm00001d", "", sceIDs)
```

Next we extract the gene IDs from the duplication data to check compatibility
```{r}
dupIDs <- duplication_info$genes$gene
head(dupIDs)
```
Here, all gene IDs are in the Zm00001ebxxxxx format; they are nuclear gene IDs from version 5. Gene IDs are thus not compatible. We will convert the IDs from the sce objects from version 4 to 5 with a conversion table.

An ID conversion table is obtained from following website after supplying a file with the IDs from sceIDs:
https://plants.ensembl.org/Zea_mays/Tools/IDMapper

```{r}
writeLines(sceIDs, here("Duplication_data", "Zma_nuclearcoding_geneIDs.txt"))
```


Read in and clean the conversion file 
- remove unneeded duplicate rows
- remove version number: e.g. Zm00001d001763.1 -> Zm00001d001763
```{r}
id_conversionTable <- read.csv(here("Duplication_data", "Zma_idConverter_nuclearCoding.txt"),
                               check.names = FALSE, 
                               stringsAsFactors = FALSE)
head(id_conversionTable)
# Clean IDs in the mapping table (Remove version suffixes like '.1' and duplicate rows)
id_conversionTable <- id_conversionTable %>%
  select(`Old stable ID`, `New stable ID`) %>%
  filter(grepl("^\\s*Zm", `Old stable ID`)) %>%
  mutate(
    Old_Cleaned = sub("\\..*", "", `Old stable ID`),
    New_Cleaned = sub("\\..*", "", `New stable ID`)
  )

head(id_conversionTable[, c("Old stable ID", "Old_Cleaned")])
nrow(id_conversionTable)
```

```{r}
# Check existence using the CLEANED IDs
final_IDmatch <- id_conversionTable %>%
  mutate(
    # Delete leading and trailing white space
    Old_Cleaned = trimws(Old_Cleaned),
    New_Cleaned = trimws(New_Cleaned),
    match_sce = Old_Cleaned %in% sceIDs,
    match_dup = New_Cleaned %in% dupIDs,
    valid = match_sce & match_dup
  )
```

```{r}
head(final_IDmatch)
unique(final_IDmatch$valid) 
```
```{r}
# Count valid matches 
cat("\nNumber of valid (mappable) genes:", sum(final_IDmatch$valid), "\n")

# Count number of sce matches
cat("\nNumber of conversion IDs matching sce IDs:", 
    sum(final_IDmatch$match_sce), "of", nrow(final_IDmatch), "\n")

# Count number of dup matches
cat("\nNumber of conversion IDs matching dup IDs:", 
    sum(final_IDmatch$match_dup), "of", nrow(final_IDmatch), "\n")
```
Of the 32927 gene IDs present in the ID conversion table, all are present in the sce object. However, 3458 are not present in the dup list
Eli: probably just genes which are not a duplicate so no problem?

## 1.5 Create paralog sce objects

Next we create new sce objects with the expression data, but only for the genes which can be linked to the duplicate information (sce object only with paralogs). V3 IDs will also be converted to V4
```{r}
# save the rows with gene IDs that are both in the dup ID list and the sce ID list
valid_matches <- final_IDmatch %>% 
  filter(valid == TRUE)
```

```{r}
# get sce ID list with only IDs which can be linked to the dup list
sce_paralog_ids <- valid_matches$Old_Cleaned

# create sce objects with only these genes
summarized_sce_paralogs <- summarized_sce[sce_paralog_ids, ]
rm(summarized_sce)
```

Confirm that order is the same
```{r}
all(rownames(summarized_sce_paralogs) == valid_matches$Old_Cleaned)
```

Convert IDs to V4
```{r}
rownames(summarized_sce_paralogs) <- valid_matches$New_Cleaned
```

The same is done for the non-summarized single cell experiment object
```{r}
sce_paralogs <- readRDS(here("chapters" ,"uncorrected_sce.rds"))
```
```{r}
sce_paralogs <- sce_paralogs[sce_paralog_ids, ]
all(rownames(sce_paralogs) == valid_matches$Old_Cleaned)
rownames(sce_paralogs) <- valid_matches$New_Cleaned
```


The file with paralog information is also filtered to contain only the gene IDs which are present in our sce objects and that could get converted with the conversion table
```{r}
# Create filtered version of duplication_info
duplication_info_filtered <- duplication_info
valid_dup_ids <- valid_matches$New_Cleaned

# Filter the genes data frame
duplication_info_filtered$genes <- duplication_info$genes %>%
  filter(trimws(gene) %in% valid_dup_ids)

# Filter the pairs data frame
duplication_info_filtered$pairs <- duplication_info$pairs %>%
  filter(trimws(dup1) %in% valid_dup_ids & trimws(dup2) %in% valid_dup_ids)
```

## 1.6 Saving

Final objects are saved for later use
```{r}
saveRDS(summarized_sce_paralogs, here("chapters", "summarized_sce_paralogs.rds"))
saveRDS(sce_paralogs, here("chapters", "sce_paralogs.rds"))
saveRDS(duplication_info, here("Duplication_data", "duplication_info.rds"))
saveRDS(duplication_info_filtered, here("Duplication_data", "duplication_info_filtered.rds"))
```


QUESTIONS
- can use other gene types besides nuclear coding?
- loss of 10k genes when constructing ID conversion table normal

```{r}
sessionInfo()
```






